---
title: "Converging sequences - CV, EBOV, RSV together"
author: "Joseph Ng"
date: "25 May 2021"
bibliography: convergent_overlap.bib
output: 
  html_document:
    df_print: kable
urlcolor: blue
---

```{r setup}
# setup and load packages
knitr::opts_chunk$set(
	echo = TRUE, message = FALSE, warning = FALSE, fig.path = "Convergence-all_plots/",  dev = "svg"
)
library(ggplot2) # for plotting
suppressPackageStartupMessages( 
  library(cowplot) # pretty ggplot
)
library(reshape2) # rearranging/'melt'-ing tables
library(plyr) # summarising (pivot-table-style) tables 
```

```{r loaddata, include=FALSE}
test_data <- read.csv('/media/josephn/My_Passport/CovidRepertoire/UseRefTrueHeavyAllCV+AllHC+AllRSV+AllEB_wo10XYFv7Mutation+Clone+AgeGroups+CorrectedSample+CombinedHC+CellType.csv',
                      stringsAsFactors = FALSE)
test_data <- test_data[grepl("IGH", test_data$Vfamily), ] 
```

# Convergent sequences

Heavy chains - Same V gene, Same J gene, Same CDR3 length, 85% CDR3 identity

CDR3s with length > 30 AAs & unproductive sequences are ignored here for simplicity.
Considered here the following all UseAsRef sequences from CV, EBOV and RSV together with Healthy controls.
Check clusters for convergent sequences from different disease conditions vs specific response.

```{r}
convg_data <- test_data[which(test_data$Num_AAs <= 30 & 
                               test_data$V.DOMAIN.Functionality == "productive"), ]
convg_data <- convg_data[which(convg_data$SampleType2 %in% c("Healthy", "Ebola",
                                                             "COVID-19", "RSV-I", "RSV-U",
                                                             "COVID-19Recovered")), ]

# first take list of IgH that binds Spike - definite ones from PDB:
pdb_ab <- list(
  read.table('known_binders/20210519ver_sabdab_SARSCoV2Abs_sequences_VJgenes.tsv', 
             sep = "\t", header = TRUE, stringsAsFactors = FALSE),
  read.table('known_binders/20210519ver_sabdab_ebolaAbs_sequences_VJgenes.tsv', 
             sep = "\t", header = TRUE, stringsAsFactors = FALSE),
  read.table('known_binders/20210519ver_sabdab_RSVAbs_sequences_VJgenes.tsv', 
             sep = "\t", header = TRUE, stringsAsFactors = FALSE)
)
pdb_ab[[1]]$Specificity <- paste0("SARS-CoV-2 ", pdb_ab[[1]]$Specificity)
pdb_ab[[2]]$Specificity <- paste0("EBOV ", pdb_ab[[2]]$Specificity)
pdb_ab[[3]]$Specificity <- paste0("RSV ", pdb_ab[[3]]$Specificity, " protein")
pdb_ab[[1]] <- pdb_ab[[1]][, 1:9]
pdb_ab <- do.call("rbind", pdb_ab)
# take only CDR3, ie removing the canonical C and W that flanks CDR3
pdb_ab$CDRH3 <- substr(pdb_ab$CDRH3, 2, nchar(pdb_ab$CDRH3)-1) 
pdb_ab$struct_id <- apply(pdb_ab[, 1:2], MARGIN = 1, paste, collapse = "|")
# pdb_ab$Specificity <- "PDB"
pdb_ab <- ddply(pdb_ab, c("CDRH3", "Specificity", "Vgene", "Jfamily"), summarise,
                  fasta_header = paste(struct_id, collapse = ","),
                  molecule_name = paste(unique(molecule_name), collapse = ","))
pdb_ab$source <- "PDB"
pdb_ab$molecule_name <- apply(pdb_ab[, c("fasta_header", "molecule_name")], 
                                MARGIN = 1, function(x) paste(x[1], x[2], sep = ": "))
pdb_ab <- pdb_ab[, -which(colnames(pdb_ab) == "fasta_header")]

# also read in known binders determined from experiments, obtained from publications
cloned_ab_pub <- c("known_binders/Dugan_et_al.txt",
                  "known_binders/Brower_et_al.txt",
                  "known_binders/Kreer_et_al.txt",
                  "known_binders/Robbiani_et_al.txt",
                  "known_binders/Gaebler_et_al.txt",
                  "known_binders/Zost_et_al.txt",
                  "known_binders/Cortjens_et_al_RSV-G.txt",
                  "known_binders/Goodwin_et_al_RSV-F_infants.txt",
                  "known_binders/Gilman_et_al_RSV-F_adults.txt",
                  "known_binders/Davis_et_al_EBOV.txt")
cloned_ab_pub <- lapply(cloned_ab_pub, read.table, stringsAsFactors=FALSE, 
                       sep = "\t", header = TRUE, comment.char = "#")
cloned_ab_pub[[1]]$CDRH3 <- substr(cloned_ab_pub[[1]]$CDRH3, 2,
                                  nchar(cloned_ab_pub[[1]]$CDRH3)-1) 
cloned_ab_pub[[2]]$CDRH3 <- substr(cloned_ab_pub[[2]]$CDRH3, 2,
                                  nchar(cloned_ab_pub[[2]]$CDRH3)-1) 
cloned_ab_pub[[3]]$CDRH3 <- substr(cloned_ab_pub[[3]]$CDRH3, 2,
                                  nchar(cloned_ab_pub[[3]]$CDRH3)-1) 
cloned_ab_pub[[7]]$CDRH3 <- substr(cloned_ab_pub[[7]]$CDRH3, 2,
                                  nchar(cloned_ab_pub[[7]]$CDRH3)-1) 
cloned_ab_pub[[8]]$CDRH3 <- substr(cloned_ab_pub[[8]]$CDRH3, 2,
                                  nchar(cloned_ab_pub[[8]]$CDRH3)-1) 
cloned_ab_pub[[9]]$CDRH3 <- substr(cloned_ab_pub[[9]]$CDRH3, 2,
                                  nchar(cloned_ab_pub[[9]]$CDRH3)-1) 
cloned_ab_pub[[4]]$Specificity <- "Spike/RBD"
cloned_ab_pub[[5]]$Specificity <- "Spike/RBD"
cloned_ab_pub[[1]]$Specificity <- paste0("SARS-CoV-2 ", cloned_ab_pub[[1]]$Specificity)
cloned_ab_pub[[2]]$Specificity <- paste0("SARS-CoV-2 ", cloned_ab_pub[[2]]$Specificity)
cloned_ab_pub[[3]]$Specificity <- paste0("SARS-CoV-2 ", cloned_ab_pub[[3]]$Specificity)
cloned_ab_pub[[4]]$Specificity <- paste0("SARS-CoV-2 ", cloned_ab_pub[[4]]$Specificity)
cloned_ab_pub[[5]]$Specificity <- paste0("SARS-CoV-2 ", cloned_ab_pub[[5]]$Specificity)
cloned_ab_pub[[6]]$Specificity <- paste0("SARS-CoV-2 ", cloned_ab_pub[[6]]$Specificity)
cloned_ab_pub[[10]]$Specificity <- paste0("EBOV ", cloned_ab_pub[[10]]$Specificity)
cloned_ab_pub[[7]]$Specificity <- "RSV G protein"
cloned_ab_pub[[8]]$Specificity <- "RSV F protein"
cloned_ab_pub[[9]]$Specificity <- "RSV F protein"
cloned_ab_pub[[1]]$source <- "Dugan_et_al"
cloned_ab_pub[[2]]$source <- "Brouwer_et_al"
cloned_ab_pub[[3]]$source <- "Kreer_et_al"
cloned_ab_pub[[4]]$source <- "Robbiani_et_al"
cloned_ab_pub[[5]]$source <- "Gaebler_et_al"
cloned_ab_pub[[6]]$source <- "Zost_et_al"
cloned_ab_pub[[7]]$source <- "Cortjens_et_al"
cloned_ab_pub[[8]]$source <- "Goodwin_et_al"
cloned_ab_pub[[9]]$source <- "Gilman_et_al"
cloned_ab_pub[[10]]$source <- "Davis_et_al"
cloned_ab_pub <- do.call("rbind", cloned_ab_pub)
cloned_ab_pub <- ddply(cloned_ab_pub, c("CDRH3", "Specificity", "Vgene", "Jfamily"),
                       summarise,
                       molecule_name = paste(sort(unique(Antibody)), collapse = ","),
                       source = paste(sort(unique(source)), collapse = ","))
cloned_ab <- rbind(pdb_ab, cloned_ab_pub)
cloned_ab <- ddply(cloned_ab, c("CDRH3", "Specificity", "Vgene", "Jfamily"), summarise,
                   molecule_name = paste(sort(unique(molecule_name)), collapse = ","),
                   source = paste(sort(unique(source)), collapse = ","))
cloned_ab$Num_AAs <- sapply(cloned_ab$CDRH3, nchar) 

# group sequences by same V and J genes
convg_data <- convg_data[, c("AA_CDR3_edited", "SampleType2", "Vgene", "Jfamily",
                             "Seq_ID", "SampleType2", "Num_AAs")]
colnames(convg_data) <- colnames(cloned_ab)
convg_data <- rbind(convg_data, cloned_ab)
convg_data <- split(convg_data, f = list(convg_data$Vgene, convg_data$Jfamily,
                                         convg_data$Num_AAs))
convg_data <- convg_data[sapply(convg_data, nrow) > 1]

```

```{r eval=FALSE}
# for each V-J combination, calculate % identity, take >= 85%
converge <- lapply(names(convg_data), function(x){
  tb <- convg_data[[x]]
  print(x)
  sims <- stringdist::stringsimmatrix(tb$CDRH3)
  sims[lower.tri(sims)] <- NA
  sims <- reshape2::melt(sims)
  sims <- sims[which(!is.na(sims[, 3])), ]
  sims <- sims[which(sims[, 1] != sims[, 2]), ]
  sims <- sims[which(sims[, 3] >= 0.85), ]
  sims[, 1] <- factor(sims[, 1], levels = 1:nrow(tb), labels = tb$molecule_name)
  sims[, 2] <- factor(sims[, 2], levels = 1:nrow(tb), labels = tb$molecule_name)
  sims
})
converge <- converge[sapply(converge, function(x) nrow(x) > 0)]
converge <- do.call("rbind", converge)

# map back metadata about the sequences
convg_data <- do.call("rbind", convg_data)
converge <- merge(converge, convg_data[, c("molecule_name", "CDRH3", "Specificity",
                                           "Vgene","Jfamily")],
                  by.x = "Var1", by.y = "molecule_name", all.x = TRUE, all.y = FALSE,
                  sort = FALSE)
colnames(converge)[4:7] <- paste0("Seq1_", colnames(converge)[4:7]) 
converge <- merge(converge, convg_data[, c("molecule_name", "CDRH3", "Specificity",
                                           "Vgene","Jfamily")],
                  by.x = "Var2", by.y = "molecule_name", all.x = TRUE, all.y = FALSE,
                  sort = FALSE)
colnames(converge)[8:11] <- paste0("Seq2_", colnames(converge)[8:11]) 
colnames(converge)[1:3] <- c("Seq2", "Seq1", "identity")

# retain only the sequences coming from different patients
converge$Seq1_PatientID <- sapply(as.character(converge$Seq1), 
                                  function(x) gsub("D[0-9]*$", "", 
                                                   unlist(strsplit(x, split = "_"))[1],
                                                   perl = TRUE))
converge$Seq2_PatientID <- sapply(as.character(converge$Seq2), 
                                  function(x) gsub("D[0-9]*$", "", 
                                                   unlist(strsplit(x, split = "_"))[1],
                                                   perl = TRUE))
converge <- converge[which(converge$Seq1_PatientID != converge$Seq2_PatientID), ]
saveRDS(converge, "ConvergeSequences_allAbPDB_graph_SameVJPatient.rds")

```

## Known binders

Combining CDRH3 from PDBs and known publications - **Total: 1813 unique CDRH3 sequences from PDB and publications**.

Evaluate CDRH3 AA identity, and construct a network connecting sequences (from data / known binders). Edges are drawn between pairs of sequences using identical rules as in connecting the convergent repertoire sequences.

```{r, fig.width=6, fig.height=4}
converge <- readRDS("ConvergeSequences_allAbPDB_graph_SameVJPatient.rds")
convg_data <- do.call("rbind", convg_data)

# process in graph object
converge_graph <- igraph::graph_from_data_frame(converge, directed = FALSE)

# the graph is very disconnected
components <- igraph::components(converge_graph)

# size distribution of components
ggplot(data.frame(components$csize)) + geom_histogram(aes(components.csize)) +
  scale_x_log10(name = "Cluster size") + cowplot::theme_cowplot() + 
  ggtitle("Size distribution of clusters") + ylab("Number of clusters")

# components with at least 10 sequences - makeup breakdown by COVID/Healthy/YFV/PDB
membership <- data.frame(components$membership)
membership$Seq_ID <- rownames(membership)
colnames(membership)[1] <- "cluster"
membership <- list(
  membership, 
  data.frame(cluster = 1:length(components$csize), cluster_size = components$csize)
)
membership <- merge(membership[[1]], membership[[2]], by = "cluster",
                    all.x = TRUE, all.y = FALSE, sort = FALSE)
converge_nodes <- list(
  converge[, which(grepl("^Seq1", colnames(converge)))],
  converge[, which(grepl("^Seq2", colnames(converge)))]
)
converge_nodes <- unique(do.call("rbind", lapply(converge_nodes, function(tb){
  colnames(tb) <- c("Seq_ID", "CDRH3", "Specificity", "Vgene", "Jfamily", "PatientID")
  tb
})))
membership <- merge(membership, converge_nodes, 
                    all.x = TRUE, all.y = FALSE, sort= FALSE)
membership$SampleType <- factor(membership$Specificity,
                                levels = c("RSV-I", "RSV-U", "COVID-19", "Healthy", "COVID-19Recovered", "Ebola",
                                           "SARS-CoV-2 Spike", "RSV F protein", "SARS-CoV-2 Spike/RBD",
                                           "SARS-CoV-2 Spike/NTD", "EBOV sGP", "RSV G protein", "EBOV GP",
                                           "EBOV NP"),
                                labels = c("Repertoire", "Repertoire", "Repertoire", "Repertoire", 
                                           "Repertoire", "Repertoire", "Binders", "Binders", "Binders",
                                           "Binders", "Binders", "Binders", "Binders", "Binders"))
# those clusters with both repertoire sequences and known binders
membership_both <- ddply(membership, "cluster", summarise,
                         both = ("Repertoire" %in% SampleType & "Binders" %in% SampleType))
membership_both <- membership_both[which(membership_both$both), "cluster"]
membership$both <- (membership$cluster %in% membership_both)

membership <- merge(membership, test_data[, c("Seq_ID", "NumInClone", "CloneGroup", "Age")],
                    by = "Seq_ID", all.x = TRUE, all.y = FALSE, sort = FALSE)
membership$SampleType2 <- membership$Specificity
saveRDS(membership, 'convergent_cluster_membership_all.rds')
```

```{r, fig.width=9, fig.height=10}
# largest clusters & V/J gene use
cluster_summary <- membership[, c("cluster", "cluster_size", "SampleType2", "Age", "PatientID", "Seq_ID",
                                  "Vgene", "Jfamily", "both")]
cluster_summary$AgeGroup <- sapply(cluster_summary$Age, function(x){
  if(x %in% c("Young", "Old")){
    if(x == "Young") return("leq50")
    if(x == "Old") return("geq60")
  } else if(is.na(x)) return(NA) else {
    x <- as.numeric(x)
    if(is.na(x)) return(NA)
    if(x <= 50) return("leq50")
    if(x >= 60) return("geq60")
  }
  return(NA)
})
cluster_summary <- cluster_summary[order(cluster_summary$cluster_size, decreasing = TRUE), ]
cluster_large <- cluster_summary[which(cluster_summary$cluster_size >= 10), ]
cluster_large <- ddply(cluster_large, c("cluster", "cluster_size", "Vgene", "Jfamily", "AgeGroup", "SampleType2"),
                       summarise, V1 = length(Seq_ID), n_patient = length(unique(PatientID)))
cluster_large$cluster <- factor(cluster_large$cluster, 
                                levels = unique(cluster_large$cluster[order(cluster_large$cluster_size, 
                                                                            decreasing = TRUE)]))

# n sequences per cluster
g1 <- ggplot(cluster_large, aes(x = cluster, y = V1, fill = SampleType2)) + geom_bar(stat = "identity") +
  cowplot::theme_cowplot() + theme(axis.text.x = element_blank()) + 
  scale_fill_manual(values = c("Healthy" = "#7FC97F", "COVID-19" = "#BEAED4", "Ebola" = "#FFFF99", 
                               "RSV-I" = "#386CB0", "RSV-U" = "#F0027F", "SARS-CoV-2 Spike" = "#99CCFF", 
                               "COVID-19Recovered" = "#FDC086", "SARS-CoV-2 Spike/RBD" = "#0000FF",
                               "RSV F protein" = "#1E2833"), name = "",
                    labels = c("CV19", "CV19-Recovered", "EBOV", "Healthy", "RSV F protein", "RSV-I", "RSV-U",
                               "SARS-CoV-2 Spike", "SARS-CoV-2 Spike/RBD")) +
  xlab("Convergent clusters") +ylab("Number of\nsequences") + ggtitle("Clusters with 10 or more sequences (n = 64)")
# present in each cohort?
g5 <- ggplot(cluster_large, aes(x = cluster, y = SampleType2)) + 
  geom_line(aes(group = cluster), color = "grey") + geom_point(aes(color = SampleType2)) +
  scale_color_manual(values = c("Healthy" = "#7FC97F", "COVID-19" = "#BEAED4", "Ebola" = "#FFFF99", 
                               "RSV-I" = "#386CB0", "RSV-U" = "#F0027F", "SARS-CoV-2 Spike" = "#99CCFF", 
                               "COVID-19Recovered" = "#FDC086", "SARS-CoV-2 Spike/RBD" = "#0000FF",
                                "SARS-CoV-2 Spike/NTD" = "#bd40ac", "RSV F protein" = "#1E2833",
                               "RSV G protein" = "#D6EAFF", "EBOV sGP" = "#99995b"), name = "",
                    labels = c("CV19", "CV19-Recovered", "EBOV", "EBOV sGP", "Healthy", "RSV F protein",
                               "RSV G protein", "RSV-I", "RSV-U", "SARS-CoV-2 Spike", "SARS-CoV-2 Spike/NTD",
                               "SARS-CoV-2 Spike/RBD")) +
  cowplot::theme_cowplot() + theme(axis.text.x = element_blank(), legend.position = "none",
                                   axis.ticks.x = element_blank()) + xlab("") +ylab("")

# Vgene usage
g2 <- ggplot(cluster_large, aes(x = cluster, y = Vgene)) + geom_tile(fill = "black") +
  cowplot::theme_cowplot() + theme(axis.text.x = element_blank(),
                                   axis.ticks.x = element_blank()) + xlab("") +ylab("V usage")
# Jfamily usage
g3 <- ggplot(cluster_large, aes(x = cluster, y = Jfamily)) + geom_tile(fill = "black") +
  cowplot::theme_cowplot() + theme(axis.text.x = element_blank(),
                                   axis.ticks.x = element_blank()) + xlab("") +ylab("J usage")

# number of donors represented (ie with sequences present) in each cluster
g4 <- ggplot(cluster_large[which(!is.na(cluster_large$AgeGroup)), ], 
             aes(x = cluster, y = n_patient, fill = AgeGroup)) + geom_bar(stat = "identity") +
  cowplot::theme_cowplot() + theme(axis.text.x = element_blank()) +
  scale_fill_manual(values = c("leq50" = "grey30", "geq60" = "grey70"),
                    label = c(expression("Age ">=60), expression("Age "<=50)), name = "") +
  xlab("Convergent clusters") +ylab("Number of\ndonors")

cowplot::plot_grid(g1, g5, g2, g3, g4, ncol = 1, axis = "lr", align = "v", rel_heights = c(2, 2, 3, 1, 1))
```

Clusters of sequences with both repertoire sequences and known binders - what are the targets?

```{r fig.width=9, fig.height=6}
both_targets <- unique(membership[membership$cluster %in% membership_both,
                                  c("Seq_ID", "cluster", "cluster_size", "SampleType2", "Specificity")])
both_targets <- ddply(both_targets, c("cluster", "cluster_size", "SampleType2", "Specificity"), nrow)
both_targets <- both_targets[order(both_targets$cluster_size, decreasing = TRUE), ]
both_targets$cluster <- factor(both_targets$cluster, levels = unique(both_targets$cluster))

# n sequences per cluster
g4 <- ggplot(unique(both_targets[, c("cluster", "Specificity", "V1")]), 
             aes(x = cluster, y = V1, fill = Specificity)) + geom_bar(stat = "identity") +
  cowplot::theme_cowplot() + theme(axis.text.x = element_blank()) + 
  scale_fill_manual(values = c("Healthy" = "#7FC97F", "COVID-19" = "#BEAED4", "Ebola" = "#FFFF99", 
                               "RSV-I" = "#386CB0", "RSV-U" = "#F0027F", "SARS-CoV-2 Spike" = "#99CCFF", 
                               "COVID-19Recovered" = "#FDC086", "SARS-CoV-2 Spike/RBD" = "#0000FF",
                                "SARS-CoV-2 Spike/NTD" = "#bd40ac", "RSV F protein" = "#1E2833",
                               "RSV G protein" = "#D6EAFF", "EBOV sGP" = "#99995b"), name = "",
                    labels = c("CV19", "CV19-Recovered", "EBOV", "EBOV sGP", "Healthy", "RSV F protein",
                               "RSV G protein", "RSV-I", "RSV-U", "SARS-CoV-2 Spike", "SARS-CoV-2 Spike/NTD",
                               "SARS-CoV-2 Spike/RBD")) +
  xlab("") +ylab("Number of\nsequences") + ggtitle("Clusters with binders (n = 52)")
# targets
g5 <- ggplot(both_targets, aes(x = cluster, y = Specificity)) + 
  geom_line(aes(group = cluster), color = "grey") + geom_point(aes(color = Specificity)) +
  scale_color_manual(values = c("Healthy" = "#7FC97F", "COVID-19" = "#BEAED4", "Ebola" = "#FFFF99", 
                               "RSV-I" = "#386CB0", "RSV-U" = "#F0027F", "SARS-CoV-2 Spike" = "#99CCFF", 
                               "COVID-19Recovered" = "#FDC086", "SARS-CoV-2 Spike/RBD" = "#0000FF",
                                "SARS-CoV-2 Spike/NTD" = "#bd40ac", "RSV F protein" = "#1E2833",
                               "RSV G protein" = "#D6EAFF", "EBOV sGP" = "#99995b"), name = "",
                    labels = c("CV19", "CV19-Recovered", "EBOV", "EBOV sGP", "Healthy", "RSV F protein",
                               "RSV G protein", "RSV-I", "RSV-U", "SARS-CoV-2 Spike", "SARS-CoV-2 Spike/NTD",
                               "SARS-CoV-2 Spike/RBD")) +
  cowplot::theme_cowplot() + theme(axis.text.x = element_blank(), legend.position = "none",
                                   axis.ticks.x = element_blank()) + xlab("Convergent clusters") +ylab("")

cowplot::plot_grid(g4, g5, ncol = 1, axis = "lr", align = "v", rel_heights = c(3,2))

```

## Example of convergent clusters

```{r }
library(visNetwork)
visualiseNetwork <- function(graph, nodes, node_annotation, cluster, node_id_col = "Seq_ID", 
                             node_color_col = "Specificity", 
                             node_colors = c("Healthy" = "#7FC97F", "COVID-19" = "#BEAED4", "Ebola" = "#FFFF99", 
                               "RSV-I" = "#386CB0", "RSV-U" = "#F0027F", "SARS-CoV-2 Spike" = "#99CCFF", 
                               "COVID-19Recovered" = "#FDC086", "SARS-CoV-2 Spike/RBD" = "#0000FF",
                                "SARS-CoV-2 Spike/NTD" = "#bd40ac", "RSV F protein" = "#1E2833",
                               "RSV G protein" = "#D6EAFF", "EBOV sGP" = "#99995b"))
{
  graphData <- visNetwork::toVisNetworkData(igraph::induced_subgraph(graph, nodes))
  graphData$edges$width <- (15 - 1) * ( graphData$edges$identity - 0.85 ) / (1 - 0.85) + 1
  graphData$edges$color <- "grey"
  graphData$nodes <- merge(graphData$nodes, node_annotation, by.x = "id", by.y = node_id_col,
                           all.x = TRUE, all.y = FALSE, sort = FALSE)
  graphData$nodes$label <- NA # no labels
  graphData$nodes$color <- graphData$nodes$Specificity
  for( i in names(node_colors) ){
    graphData$nodes$color <- replace(graphData$nodes$color, 
                                     which(graphData$nodes[, node_color_col] == i),
                                     node_colors[i])
  }
  lnodes <- data.frame(node_colors)
  colnames(lnodes)[1] <- "color"
  lnodes$label <- rownames(lnodes)
  lnodes$color.border <- "white"
  lnodes$title <- ""
  lnodes$id <- 1:nrow(lnodes)
  visNetwork(nodes = graphData$nodes, edges = graphData$edges, main = paste0("Cluster ", cluster)) %>%
    visLegend(addNodes = lnodes, useGroups = FALSE) %>% visIgraphLayout()
}
visualiseNetwork(converge_graph, membership[membership$cluster == 1833, 1], cluster = 1833,
                 node_annotation = converge_nodes)
visualiseNetwork(converge_graph, membership[membership$cluster == 1058, 1], cluster = 1058,
                 node_annotation = converge_nodes)
visualiseNetwork(converge_graph, membership[membership$cluster == 932, 1], cluster = 932,
                 node_annotation = converge_nodes)
visualiseNetwork(converge_graph, membership[membership$cluster == 2691, 1], cluster = 2691,
                 node_annotation = converge_nodes)
visualiseNetwork(converge_graph, membership[membership$cluster == 2159, 1], cluster = 2159,
                 node_annotation = converge_nodes)
visualiseNetwork(converge_graph, membership[membership$cluster == 30, 1], cluster = 30,
                 node_annotation = converge_nodes)
visualiseNetwork(converge_graph, membership[membership$cluster == 1630, 1], cluster = 1630,
                 node_annotation = converge_nodes)
visualiseNetwork(converge_graph, membership[membership$cluster == 1718, 1], cluster = 1718,
                 node_annotation = converge_nodes)

```

CDRH3 AA Sequence logos of these clusters:

```{r fig.width=5, fig.height=5}
seqs <- membership[membership$cluster %in% c("1833", "1058", "932", "2691", "2159", "30", "1630", "1718"),
                   c("Seq_ID", "cluster", "CDRH3")]
seqs <- merge(seqs, ddply(seqs, "cluster", nrow), by = "cluster")
seqs$cluster <- paste0("Cluster ", seqs$cluster, "\n(n = ", seqs$V1, ")")

library(ggseqlogo)
cowplot::plot_grid(
  plotlist = lapply(c("1833", "1058", "932", "2691", "2159", "30", "1630", "1718"), function(x){
    ggseqlogo(seqs[which(grepl(paste0("Cluster ", x), seqs$cluster)), "CDRH3"]) +
        ylab(unique(seqs[which(grepl(paste0("Cluster ", x), seqs$cluster)), "cluster"])) +
        theme_cowplot() + theme(legend.position = "none", axis.text = element_blank(), axis.ticks = element_blank(),
                                axis.line = element_blank(), axis.title.y = element_text(angle = 0, vjust = 0.5))
  }), align = "v", axis = "lr", ncol = 1
)
```

## Analysis

### Expansion

```{r, fig.width=9, fig.height=4}
membership$large <- (membership$cluster_size >= 10)
membership$SampleType2 <- replace(membership$Specificity,
                                  which(membership$Specificity == "COVID-19Recovered"),
                                  "COVID-19")
test_data$SampleType3 <- replace(test_data$SampleType2,
                                  which(test_data$SampleType2 == "COVID-19Recovered"),
                                  "COVID-19")
expansion <- ddply(membership, c("SampleType2", "large", "CloneGroup"), nrow)
expansion <- expansion[which(expansion$SampleType2 %in% c("COVID-19", "Ebola", "RSV-I", "RSV-U", "Healthy")), ]
expansion <- merge(
  expansion, ddply(membership, c("SampleType2", "large"), nrow), all.x = TRUE, all.y = FALSE,
  by = c("SampleType2", "large"), sort = FALSE
)
expansion$perc <- expansion[, 4] / expansion[, 5]
colnames(expansion) <- c("SampleType2", "large", "CloneGroup", "n", "total", "perc")
expansion$type <- "convergent"
expansion <- list(
  expansion, 
  merge(
    ddply(test_data[!test_data$Seq_ID %in% membership$Seq_ID, ], 
          c("SampleType3", "CloneGroup"), nrow), 
    ddply(test_data[!test_data$Seq_ID %in% membership$Seq_ID, ], 
          "SampleType2", nrow), all.x = TRUE, all.y = FALSE,
    by.x = "SampleType3", by.y = "SampleType2", sort = FALSE
  )
)
expansion[[2]]$perc <- expansion[[2]][, 3] / expansion[[2]][, 4]
colnames(expansion[[2]]) <- c("SampleType2", "CloneGroup", "n", "total", "perc")
expansion[[2]] <- expansion[[2]][which(expansion[[2]]$SampleType2 %in% c("COVID-19", "Ebola", "RSV-I", "RSV-U", "Healthy")), ]
expansion[[2]]$type <- "other sequences"
expansion[[2]]$large <- "others"
expansion[[2]] <- expansion[[2]][, c("SampleType2", "large", "CloneGroup", "n", "total", "perc", "type")]
expansion <- do.call("rbind", expansion)
expansion$CloneGroup <- factor(expansion$CloneGroup,
                               levels = c("Unique", "2", "3", "4&5", "6to9", ">10"),
                               labels = c("Unique", "2", "3", "4-5", "6-9", 
                                          "10 or more"))
expansion$large <- factor(expansion$large, levels = c(TRUE, FALSE, "others"),
                          labels = c("convergent\n(10 or more\nsequences)", "convergent\n(<10 sequences)",
                                     "other\nsequences"))
ggplot(expansion, aes(y = CloneGroup, x = perc)) + 
  geom_bar(stat = "identity", position = position_dodge(), width = 0.5) +
  scale_x_continuous(labels = scales::percent, name = "% sequences") +
  ylab("Number of sequences in clone") + cowplot::theme_cowplot() + 
  facet_grid(large ~ SampleType2) + theme(strip.text = element_text(size = 9),
                                          axis.text.x = element_text(angle = 45, hjust = 1))

```

Another way of presenting this, treating NumInClone as a continuous variable:

```{r, fig.width=8, fig.height=4}
expansion_cont <- membership[membership$SampleType2 %in% c("COVID-19", "Ebola", "RSV-I", "RSV-U", "Healthy"),
                             c("Seq_ID", "large", "NumInClone", "SampleType2")]
expansion_cont <- list(
  expansion_cont,
  test_data[which(!test_data$Seq_ID %in% expansion_cont$Seq_ID & 
              test_data$SampleType2 %in% c("COVID-19", "Ebola", "RSV-I", "RSV-U", "Healthy")), 
            c("Seq_ID", "SampleType", "NumInClone", "SampleType2")]
)
expansion_cont[[2]][, 2] <- "others"
colnames(expansion_cont[[2]]) <- c("Seq_ID", "large", "NumInClone", "SampleType2")
expansion_cont <- do.call("rbind", expansion_cont)
expansion_cont$large <- factor(expansion_cont$large, levels = c(TRUE, FALSE, "others"),
                          labels = c("convergent\n(10 or moresequences)", "convergent\n(<10 sequences)",
                                     "other sequences"))

ggplot(expansion_cont, aes(x = large, y = NumInClone)) + geom_boxplot(outlier.shape = NA) + 
  facet_grid(~ SampleType2, drop = TRUE) + 
  scale_y_log10(limits = c(1, 500), name = "Number of sequences\nin clone") + cowplot::theme_cowplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + xlab("")

# cant make the ggpubr::compare_means add brackets & p-value * properly so here are the data
ggpubr::compare_means(NumInClone ~ large, group.by = "SampleType2", 
                      data = expansion_cont, method = "wilcox", ref.group = "other sequences")
```

# References
