---
title: "Class-switching analysis of Repertoires"
author: "Joseph Ng"
date: "8 April 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "svg", fig.path = "CSR_plots/")
library(plyr)
library(ggplot2)
```

Analysis of CSR by generating pruned "arborescence graphs" from parismony-based reconstructed lineage trees.

Arborescence graphs (definition: there is exactly one directed path from the root to every other nodes in the graph) were generated respecting the rules of CSR such that only paths between nodes (sequences) of either (a) the same isotype or (b) an isotype downstream are retained.

This allows for analysis of both the number and timing of CSR events within lineages.

```{r loaddata, include=FALSE}
test_data <- read.table('/media/josephn/My_Passport/CovidRepertoire/UseRefTrueHeavyAllCV+AllHC+AllRSV+AllEB_wo10XYFv7Mutation+Clone+AgeGroups+CorrectedSample+CombinedHC+CellType.csv', header = TRUE, sep = ",",
                      stringsAsFactors = FALSE, comment.char = "")
test_data <- test_data[ which(grepl("^IGH", test_data$Vfamily)), ]
test_data$CloneID <- apply(test_data[, c("PatientID", "CloneID")], MARGIN = 1,
                           function(x) paste(x[1], x[2], sep = "_"))
test_data$CloneID <- gsub(" ", "", test_data$CloneID)

# further separate Ebola by ethnicity and group all COVID-19 (timepoints + FU) as COVID-19
test_data$SampleType3 <- test_data$SampleType2
test_data$SampleType3 <- replace(test_data$SampleType3,
                                 which(grepl("^CV", test_data$PatientID) & grepl("^FU", test_data$SampleName)),
                                 "CV19")
test_data$SampleType3 <- replace(test_data$SampleType3,
                                 which(test_data$PatientID %in% c("EB215", "EB216", "EB217")),
                                 "EBOV - White")
test_data$SampleType3 <- replace(test_data$SampleType3,
                                 which(test_data$PatientID %in% c("E2", "E50", "E14", "E82", "E15", 
                                                                  "E43", "E46", "E70", "E75")),
                                 "EBOV - West African")
test_data$SampleType3 <- replace(test_data$SampleType3,
                                 which(test_data$PatientID %in% c("E2", "E50", "E14", "E82", "E15", 
                                                                  "E43", "E46", "E70", "E75")),
                                 "EBOV - West African")

# set up age group
test_data$AgeGroup <- test_data$Age
test_data$AgeGroup[ which(as.numeric(test_data$Age) <= 50 )] <-  "<=50"
test_data$AgeGroup[ which(as.numeric(test_data$Age) >= 60 )] <-  ">=60"
test_data$AgeGroup[ which(as.numeric(test_data$Age) > 50 & as.numeric(test_data$Age < 60))] <- NA
#test_data$AgeGroup[ which(as.numeric(test_data$Age) > 65 )] <-  "Old"
test_data$AgeGroup[ which(test_data$Age == "nd")] <- NA
test_data$AgeGroup[ which(test_data$Age == "Young")] <- "<=50"
test_data$AgeGroup[ which(test_data$Age == "Old")] <- ">=60"
#colnames(test_data)[ncol(test_data)] <- "AgeGroup"
```

# Lineage

## Number of class switching events

```{r}
tree_stats <- list.files(path = '/media/josephn/My_Passport/CovidRepertoire/tree_stats', 
                         pattern = "rds$", full.names = TRUE)
tree_stats <- tree_stats[which(!grepl("YF189|YF191|YF192|YF200|YF203", tree_stats))]
tree_stats <- lapply(tree_stats, function(x){
  ts <- readRDS(x)
  sample_name <- basename(x)
  sample_name <- unlist(strsplit(sample_name, split = "_"))[1]
  ts <- lapply(names(ts), function(y){
    if( length(ts[[y]]) == 0 ) return(NULL)
    tb <- ts[[y]]$csr_events
    if( is.null(tb) ) return(NULL)
    if( nrow(tb) == 0 ) return(NULL)
    tb$CloneID <- y
    tb
  })
  ts <- ts[which(sapply(ts, function(z) !is.null(z)))]
  if(length(ts) > 0){
    ts <- do.call("rbind", ts)
    ts$PatientId <- sample_name
    ts
  } else return(data.frame())
})
tree_stats <- do.call("rbind", tree_stats)
tree_stats <- merge(tree_stats, 
                    unique(test_data[, c("CloneID", "AgeGroup", "NumInClone", 
                                         "CloneGroup")]), all.x = TRUE, all.y = FALSE,
                    sort = FALSE, by.x = "CloneID", by.y = "CloneID")
#____________________________
# add back SampleType
# rules:
# (1) Ebola - as is
# (2) COVID - COVID TimePoints + FU
# (3) Healthy - as is
# (4) RSV - (a) Both sequences are from D0: Healthy; (b) At least one sequence NOT from D0: RSV-I / RSV-U as appropriate
# (5) YFV - (a) Both sequences are from D0: Healthy; (b) At least one sequence NOT from D0: YFVD28
#____________________________
tree_stats$SampleType2 <- apply(tree_stats[, c("startLabel", "endLabel", "PatientId")], MARGIN = 1, function(x){
  if(grepl("^CV", x[3])) return("COVID-19")
  if(x[3] %in% c("EB215", "EB216", "EB217")) return("EBOV - White")
  if(x[3] %in% c("E2", "E50", "E14", "E82", "E15", "E43", "E46", "E70", "E75")) return("EBOV - West African")
  if(grepl("^HB", x[3])) return("Healthy")
  if(x[3] == "YR001"){
    seq_names <- c(unlist(strsplit(x[1], split = "_"))[1], unlist(strsplit(x[2], split = "_"))[1])
    if(grepl("D0", seq_names[1]) & grepl("D0", seq_names[2])) return("Healthy") 
    else if(grepl("D0", seq_names[1]) & grepl("D10", seq_names[2])) return("RSV-I")
    else return(NA)
  }
  if(x[3] %in% c("YR004", "YR008", "EA003", "EA005", "EA008")){
    seq_names <- c(unlist(strsplit(x[1], split = "_"))[1], unlist(strsplit(x[2], split = "_"))[1])
    if(grepl("D0", seq_names[1]) & grepl("D0", seq_names[2])) return("Healthy") else return("RSV-I")
  }
  if(x[3] %in% c("YR002", "YR003", "YR006", "EA014", "EA022", "EA024")){
    seq_names <- c(unlist(strsplit(x[1], split = "_"))[1], unlist(strsplit(x[2], split = "_"))[1])
    if(grepl("D0", seq_names[1]) & grepl("D0", seq_names[2])) return("Healthy") else return("RSV-U")
  }
  if(x[3] %in% c("YF193", "YF199", "YF207")){
    seq_names <- c(unlist(strsplit(x[1], split = "_"))[1], unlist(strsplit(x[2], split = "_"))[1])
    if(grepl("D0", seq_names[1]) & grepl("D0", seq_names[2])) return("Healthy") else return("YFVD28")
  }
})
summary(factor(tree_stats$SampleType2))

```

Here too few YFVD28. Here ignore them for this analysis. Moreover the NA (i.e. 'Infected' in the original data table - RSV Infected multiple-timepoints from 1 patient) are also omitted.

```{r}
tree_stats <- tree_stats[which(tree_stats$SampleType2 != "YFVD28"), ]
# frequency & distFromGermline of CSR
tree_bySampleType <- ddply(
  tree_stats[which(!is.na(tree_stats$SampleType2)), ], c("SampleType2", "AgeGroup", "PatientId", "CloneGroup"),
  summarise,
  nEvent = length(unique(CloneID)),
  meanDist_from_germline = mean(distFromGermline, na.rm = TRUE)
)
tree_bySampleType <- merge(tree_bySampleType,
                         ddply(test_data[which(test_data$NumInClone >= 3 & test_data$SampleType2 != "Infected"),], 
                               c("PatientID"), summarise, 
                               nClone = length(unique(CloneID))),
                         by.x = "PatientId", by.y = "PatientID", sort = FALSE)
tree_bySampleType$nClone_events <- tree_bySampleType$nEvent / tree_bySampleType$nClone
tree_bySampleType$AgeGroup <- factor(tree_bySampleType$AgeGroup,
                                   levels = c("<=50", ">=60"))
tree_bySampleType$SampleType2 <- factor(tree_bySampleType$SampleType2,
                                      levels = c("Healthy", "COVID-19", "EBOV - White", "EBOV - West African",
                                                 "RSV-I", "RSV-U", "YFVD28"),
                                      labels = c("Healthy", "CV19", "EBOV", "EBOV",
                                                 "RSV-I", "RSV-U", "YFVD28"))
tree_bySampleType$CloneGroup <- factor(tree_bySampleType$CloneGroup,
                                     levels = c("Unique", "2", "3", "4&5",
                                                "6to9", ">10"),
                                     labels = c("Unique", "2", "3", "4 & 5",
                                                "6 to 9", ">10"))
```

```{r, fig.width=6, fig.height=6}
# evaluate statistical significance in the occurrence of CSR events
nEvent_test <- lapply(unique(tree_bySampleType$CloneGroup), function(x){
  require(multcomp)
  o <- lapply(1, function(y){#c("Young", "Middle", "Old"), function(y){
    lm_tb <- tree_bySampleType[which(tree_bySampleType$CloneGroup == x), ]
    colnames(lm_tb)[ which(colnames(lm_tb) == "SampleType2") ] <- "xxx"
    
    # One-way ANOVA
    if(nrow(lm_tb) > 0 & length(unique(lm_tb$xxx)) > 1 &
       sum(grepl("^Healthy", lm_tb$xxx)) > 0){
      onewayanova <- try( suppressWarnings(
        aov(
          as.formula( paste("nClone_events", "xxx", sep = " ~ ")), data = lm_tb
        )), silent = TRUE)
      if( "aov" %in% class(onewayanova) ){
        # Do Post-hoc test (Dunnett)
        p <- summary(onewayanova)[[1]][1,5]
        if( is.nan(p) ) posthoc <- NA else {
          posthoc <- glht(onewayanova, linfct = mcp(xxx = "Dunnett"))
          posthoc <- summary(posthoc)
          signif_comparison <- 1:length(posthoc$test$pvalues)#which(posthoc$test$pvalues < 0.05)
          if( length(signif_comparison) > 0){
            posthoc <- lapply(signif_comparison, function(y){
              data.frame(variable = x,
                         SampleType2 = unlist(strsplit(names(posthoc$test$coefficients)[y], split = " - "))[1],
                         pval = round(posthoc$test$pvalues[y], 4))
            })
            posthoc <- do.call("rbind", posthoc)
          } else posthoc <- data.frame()
        }
      } else return(data.frame())
    } else return(data.frame())
  })
  do.call("rbind", o)
})
nEvent_test <- do.call("rbind", nEvent_test)
rownames(nEvent_test) <- NULL

# put statistical test results back into the summary table
tree_bySampleType <- merge(tree_bySampleType, nEvent_test, by.x = c("CloneGroup", "SampleType2"),
                         by.y = c("variable", "SampleType2"), sort = FALSE, all.x = TRUE)
tree_bySampleType$sig <- (tree_bySampleType$pval < 0.05)
tree_bySampleType$sig[is.na(tree_bySampleType$sig)] <- FALSE

# plot
ggplot(tree_bySampleType[!tree_bySampleType$CloneGroup %in% c("Unique", "2"), ], 
       aes(x = SampleType2, y = nClone_events, colour = sig)) + geom_boxplot() +
  geom_point(position = position_jitter(width = 0.2)) + 
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.2), 
                     name = "% clones with CSR events") +
  scale_colour_manual(values = c("FALSE" = "grey", "TRUE" = "red"), name = "p < 0.05") +
  cowplot::theme_cowplot() + xlab("") + theme(legend.position = "bottom") +
  facet_wrap(~ CloneGroup, ncol = 1, strip.position = "right")
```

Split by age:

```{r, fig.width=7, fig.height=6}
ggplot(tree_bySampleType[!tree_bySampleType$CloneGroup %in% c("Unique", "2"), ], 
       aes(x = SampleType2, y = nClone_events, colour = AgeGroup)) + geom_boxplot() +
    geom_point(position = position_dodge(width = 0.7)) + 
    scale_y_continuous(labels = scales::percent, limits = c(0, 0.2), 
                       name = "% clones with CSR events") +
    cowplot::theme_cowplot() + xlab("") + theme(legend.position = "bottom") +
    facet_wrap(~ CloneGroup, ncol = 1, strip.position = "right")

```

## Distance-from-germline of class switching events

Parsing arborescence graphs for CSR events and getting distances from germline - this estimates when ('timing') a switch occurs, relative to the lineage.

```{r, fig.width=8, fig.height=6}
# frequency & distFromGermline of CSR by agegroup/healthy
tree_bySampleType <- ddply(
  tree_stats[which(!is.na(tree_stats$SampleType2)), ], c("SampleType2", "startIsotype", "endIsotype"),
  summarise,
  nEvent = length(unique(CloneID)),
  meanDist_from_germline = mean(distFromGermline, na.rm = TRUE)
)
tree_bySampleType <- merge(tree_bySampleType,
                         ddply(test_data[which(test_data$NumInClone >= 3 & test_data$SampleType2 != "Infected"),], 
                               c("SampleType3"), summarise, 
                               nClone = length(unique(CloneID))),
                         by.x = c("SampleType2"), by.y = c("SampleType3"))
tree_bySampleType$nClone_events <- tree_bySampleType$nEvent / tree_bySampleType$nClone
tree_bySampleType <- tree_bySampleType[which(grepl("Ig[MDE]|IgA[12]|IgG[1-4]",
                                               tree_bySampleType$startIsotype) &
                                           grepl("Ig[MDE]|IgA[12]|IgG[1-4]",
                                               tree_bySampleType$endIsotype)), ]
tree_bySampleType$startIsotype <- factor(tree_bySampleType$startIsotype,
                                       levels = c("IgM", "IgG3", "IgG1", "IgA1",
                                                  "IgG2", "IgG4", "IgE", "IgA2"),
                                       labels = c("M", "G3", "G1", "A1", "G2",
                                                  "G4", "E", "A2"))
tree_bySampleType$endIsotype <- factor(tree_bySampleType$endIsotype,
                                     levels = c("IgM", "IgG3", "IgG1", "IgA1",
                                                "IgG2", "IgG4", "IgE", "IgA2"),
                                       labels = c("M", "G3", "G1", "A1", "G2",
                                                  "G4", "E", "A2"))
#tree_bySampleType$AgeGroup <- factor(tree_bySampleType$AgeGroup,
#                                   levels = c("?", "Young", "Middle", "Old"))
tree_bySampleType$SampleType2 <- factor(tree_bySampleType$SampleType2,
                                          levels = c("Healthy", "COVID-19", "EBOV - White",
                                                     "RSV-I", "RSV-U", "EBOV - West African"),
                                        labels = c("Healthy", "CV19", "EBOV - White", "RSV-I",
                                                   "RSV-U", "EBOV - West African"))

#plot
ggplot(tree_bySampleType[which(tree_bySampleType$SampleType2 %in% 
                               c("Healthy", "CV19", "EBOV - White", "EBOV - West African", "RSV-I", "RSV-U") &
                               tree_bySampleType$endIsotype != "E"), ],#[which(!tree_bySampleType$SampleType %in% 
                      #         c("YFV Vaccine")), ], 
       aes(x = endIsotype, y = startIsotype, 
           fill = meanDist_from_germline, size = nClone_events)) +
  geom_point(pch = 21) + cowplot::theme_cowplot() + 
  scale_size_continuous(range = c(0, 10), labels = scales::percent,
                        name = "% clones with\nCSR events") +
  scale_fill_viridis_c(limits = c(0, 0.2), 
                       name = "CSR timing\n(mean distance\nfrom germline)") +
  xlab("To") + ylab("From") + facet_wrap( ~ SampleType2)
```

**Statistical test for significance**: One-way ANOVA followed by Dunnett post-hoc comparison against the Healthy cohort.

## Statistical test: CSR timing

```{r}
# Individual patients:
# frequency & distFromGermline of CSR by agegroup/healthy
tree_byPatient <- ddply(
  tree_stats[which(!is.na(tree_stats$SampleType2)), ], 
  c("SampleType2", "PatientId", "startIsotype", "endIsotype"),
  summarise,
  nEvent = length(unique(CloneID)),
  meanDist_from_germline = mean(distFromGermline, na.rm = TRUE)
)
tree_byPatient <- merge(tree_byPatient,
                         ddply(test_data[which(test_data$NumInClone >= 3 & test_data$SampleType2 != "Infected"),], 
                               c("PatientID"), summarise, 
                               nClone = length(unique(CloneID))),
                        by.x = "PatientId", by.y = "PatientID", sort = FALSE)
tree_byPatient$nClone_event <- tree_byPatient$nEvent / tree_byPatient$nClone
tree_byPatient <- tree_byPatient[which(grepl("Ig[MDE]|IgA[12]|IgG[1-4]",
                                               tree_byPatient$startIsotype) &
                                           grepl("Ig[MDE]|IgA[12]|IgG[1-4]",
                                               tree_byPatient$endIsotype)), ]
tree_byPatient$startIsotype <- factor(tree_byPatient$startIsotype,
                                       levels = c("IgM", "IgG3", "IgG1", "IgA1",
                                                  "IgG2", "IgG4", "IgE", "IgA2"),
                                       labels = c("M", "G3", "G1", "A1", "G2",
                                                  "G4", "E", "A2"))
tree_byPatient$endIsotype <- factor(tree_byPatient$endIsotype,
                                     levels = c("IgM", "IgG3", "IgG1", "IgA1",
                                                "IgG2", "IgG4", "IgE", "IgA2"),
                                       labels = c("M", "G3", "G1", "A1", "G2",
                                                  "G4", "E", "A2"))
#tree_byPatient$AgeGroup <- factor(tree_byPatient$AgeGroup,
#                                  levels = c("?", "Young", "Middle", "Old"))
tree_byPatient$SampleType2 <- factor(tree_byPatient$SampleType2,
                                    levels = c("Healthy", "COVID-19", "EBOV - West African", "EBOV - White",
                                               "RSV-I", "RSV-U", "YFVD28"),
                                    labels = c("Healthy", "CV19", "EBOV (West African)", "EBOV (White)",
                                               "RSV-I", "RSV-U", "YFVD28"))

# evaluate statistical significance
timing_test <- apply(combn(c("M", "G3", "G1", "A1", "G2", "G4", "E", "A2"), 2), 
                     MARGIN = 2, function(x){
  o <- lapply(1, function(y){#c("Young", "Middle", "Old"), function(y){
    lm_tb <- tree_byPatient[which(tree_byPatient$startIsotype == x[1] &
                                    tree_byPatient$endIsotype == x[2]), ]
#                                    tree_byPatient$AgeGroup == y), ]
    colnames(lm_tb)[which(colnames(lm_tb) == "SampleType2")] <- "xxx"
    
    # One-way ANOVA
    if(nrow(lm_tb) > 0 & length(unique(lm_tb$xxx)) > 1 &
       sum(grepl("^Healthy", lm_tb$xxx)) > 0){
      onewayanova <- try( suppressWarnings(
        aov(
          as.formula( paste("meanDist_from_germline", "xxx", sep = " ~ ")), data = lm_tb
        )), silent = TRUE)
      if( "aov" %in% class(onewayanova) ){
        # Do Post-hoc test (Dunnett)
        p <- summary(onewayanova)[[1]][1,5]
        if( is.null(p) ) posthoc <- data.frame() else {
          posthoc <- glht(onewayanova, linfct = mcp(xxx = "Dunnett"))
          posthoc <- summary(posthoc)
          signif_comparison <- 1:length(posthoc$test$pvalues)#which(posthoc$test$pvalues < 0.05)
          if( length(signif_comparison) > 0){
            posthoc <- lapply(signif_comparison, function(y){
              data.frame(startIsotype = x[1], endIsotype = x[2],
                         SampleType2 = unlist(strsplit(names(posthoc$test$coefficients)[y], split = " - "))[1],
                         pval = round(posthoc$test$pvalues[y], 4))
            })
            posthoc <- do.call("rbind", posthoc)
          } else posthoc <- data.frame()
        }
      } else return(data.frame())
    } else return(data.frame())
  })
  do.call("rbind", o)
})
timing_test <- do.call("rbind", timing_test)
rownames(timing_test) <- NULL
timing_test <- timing_test[timing_test$pval < 0.05, ]
timing_test
```

```{r, fig.width=7, fig.height=5}
# plot
tree_byPatient$CSR <- apply(tree_byPatient[, c("startIsotype", "endIsotype")],
                            MARGIN = 1, function(x) paste(x[1], x[2], sep = " to "))
timing_test$CSR <- apply(timing_test[, c("startIsotype", "endIsotype")],
                         MARGIN = 1, function(x) paste(x[1], x[2], sep = " to "))
tree_byPatient <- merge(tree_byPatient, 
                        timing_test[, c("SampleType2", "CSR", "pval")],
                        by = c("SampleType2", "CSR"), all.x = TRUE,
                        sort = FALSE)
colnames(tree_byPatient)[ncol(tree_byPatient)] <- "p_timing"
tree_byPatient$signif_timing <- (tree_byPatient$p_timing < 0.05)
tree_byPatient[is.na(tree_byPatient$signif_timing), "signif_timing"] <- FALSE

# plot those with significant differences in at least 1 sample type
ggplot(tree_byPatient[which(tree_byPatient$CSR %in% timing_test$CSR),],
       aes(x = SampleType2, y = meanDist_from_germline, colour = signif_timing)) +
  geom_boxplot() + geom_point() + cowplot::theme_cowplot() +
  facet_grid(~ CSR) + xlab("") + ylab("mean distance from germline") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
  scale_colour_manual(values = c("TRUE" = "red", "FALSE" = "grey"), name = "p < 0.05")
```

Split by age group. Model $Y = \text{agegroup}\ *\ \text{SampleType} $ and Tukey's post-hoc test.

```{r}
# Individual patients:
# frequency & distFromGermline of CSR by agegroup/healthy
tree_byPatient <- ddply(
  tree_stats[which(!is.na(tree_stats$SampleType2)), ], 
  c("SampleType2", "AgeGroup", "PatientId", "startIsotype", "endIsotype"),
  summarise,
  nEvent = length(unique(CloneID)),
  meanDist_from_germline = mean(distFromGermline, na.rm = TRUE)
)
tree_byPatient <- merge(tree_byPatient,
                         ddply(test_data[which(test_data$NumInClone >= 3 & test_data$SampleType2 != "Infected"),], 
                               c("PatientID"), summarise, 
                               nClone = length(unique(CloneID))),
                        by.x = "PatientId", by.y = "PatientID", sort = FALSE)
tree_byPatient$nClone_event <- tree_byPatient$nEvent / tree_byPatient$nClone
tree_byPatient <- tree_byPatient[which(grepl("Ig[MDE]|IgA[12]|IgG[1-4]",
                                               tree_byPatient$startIsotype) &
                                           grepl("Ig[MDE]|IgA[12]|IgG[1-4]",
                                               tree_byPatient$endIsotype)), ]
tree_byPatient$startIsotype <- factor(tree_byPatient$startIsotype,
                                       levels = c("IgM", "IgG3", "IgG1", "IgA1",
                                                  "IgG2", "IgG4", "IgE", "IgA2"),
                                       labels = c("M", "G3", "G1", "A1", "G2",
                                                  "G4", "E", "A2"))
tree_byPatient$endIsotype <- factor(tree_byPatient$endIsotype,
                                     levels = c("IgM", "IgG3", "IgG1", "IgA1",
                                                "IgG2", "IgG4", "IgE", "IgA2"),
                                       labels = c("M", "G3", "G1", "A1", "G2",
                                                  "G4", "E", "A2"))
#tree_byPatient$AgeGroup <- factor(tree_byPatient$AgeGroup,
#                                  levels = c("?", "Young", "Middle", "Old"))
tree_byPatient$SampleType2 <- factor(tree_byPatient$SampleType2,
                                    levels = c("Healthy", "COVID-19", "EBOV - West African", "EBOV - White",
                                               "RSV-I", "RSV-U", "YFVD28"),
                                    labels = c("Healthy", "CV19", "EBOV (West African)", "EBOV (White)",
                                               "RSV-I", "RSV-U", "YFVD28"))

# evaluate statistical significance
timing_test <- apply(combn(c("M", "G3", "G1", "A1", "G2", "G4", "E", "A2"), 2), 
                     MARGIN = 2, function(x){
  o <- lapply(1, function(y){#c("Young", "Middle", "Old"), function(y){
    lm_tb <- tree_byPatient[which(tree_byPatient$startIsotype == x[1] &
                                    tree_byPatient$endIsotype == x[2]), ]
#                                    tree_byPatient$AgeGroup == y), ]
    # colnames(lm_tb)[which(colnames(lm_tb) == "SampleType2")] <- "xxx"
    lm_tb$AgeSample <- interaction(lm_tb$AgeGroup, lm_tb$SampleType2)
    # Two-way ANOVA
    if(nrow(lm_tb) > 0 & length(unique(lm_tb$SampleType2)) > 1 &
       sum(grepl("^Healthy", lm_tb$SampleType2)) > 0){
      twowayanova <- try( suppressWarnings(
        aov(
          as.formula( paste("meanDist_from_germline ~ AgeSample - 1")), data = lm_tb
        )), silent = TRUE)
      if( "aov" %in% class(twowayanova) ){
        # Do Post-hoc test (Dunnett)
        p <- summary(twowayanova)[[1]][1,5]
        if( is.null(p) ) posthoc <- data.frame() else {
          Tukey <- table(lm_tb$AgeSample)
          Tukey <- contrMat(Tukey[Tukey > 0], "Tukey")
          # comparison of either sample age group & any sample type vs Healthy, or
          # same sample type & different age groups
          Tukey <- list(
            Tukey[which(grepl("Healthy", rownames(Tukey)) & stringr::str_count(rownames(Tukey), "<=50") != 1), ],
            Tukey[which(sapply(stringr::str_extract_all(rownames(Tukey), "\\.([A-Za-z\\-]*)"), 
                               function(z) z[1] ==z[2])), ]
          )
          Tukey <- do.call("rbind", Tukey)
          posthoc <- glht(twowayanova, linfct = Tukey)
          posthoc <- summary(posthoc)
          signif_comparison <- 1:length(posthoc$test$pvalues)#which(posthoc$test$pvalues < 0.05)
          if( length(signif_comparison) > 0){
            posthoc <- lapply(signif_comparison, function(y){
              data.frame(startIsotype = x[1], endIsotype = x[2],
                         group1 = unlist(strsplit(names(posthoc$test$coefficients)[y], split = " - "))[1],
                         group2 = unlist(strsplit(names(posthoc$test$coefficients)[y], split = " - "))[2],
                         pval = round(posthoc$test$pvalues[y], 4))
            })
            posthoc <- do.call("rbind", posthoc)
          } else posthoc <- data.frame()
        }
      } else return(data.frame())
    } else return(data.frame())
  })
  do.call("rbind", o)
})
timing_test <- do.call("rbind", timing_test)
rownames(timing_test) <- NULL
#timing_test <- timing_test
timing_test[timing_test$pval < 0.05, ]
```

```{r, fig.width=8, fig.height=8}
# plot
tree_byPatient$CSR <- apply(tree_byPatient[, c("startIsotype", "endIsotype")],
                            MARGIN = 1, function(x) paste(x[1], x[2], sep = " to "))
timing_test$CSR <- apply(timing_test[, c("startIsotype", "endIsotype")],
                         MARGIN = 1, function(x) paste(x[1], x[2], sep = " to "))

# plot those with significant differences in at least 1 sample type
cowplot::plot_grid(
  ggplot(tree_byPatient[which(tree_byPatient$CSR %in% timing_test[timing_test$pval < 0.05, "CSR"] & 
                              !is.na(tree_byPatient$AgeGroup) &
                              tree_byPatient$SampleType2 %in% c("Healthy", "CV19", "EBOV (White)")),],
       aes(x = SampleType2, y = meanDist_from_germline, colour = AgeGroup)) +
  geom_boxplot() + geom_point(position = position_jitterdodge()) + cowplot::theme_cowplot() +
  facet_grid(~ CSR) + xlab("") + ylab("mean distance from germline") + ylim(0, 0.28) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)),
  ggplot(tree_byPatient[which(tree_byPatient$CSR %in% timing_test[timing_test$pval < 0.05, "CSR"] & 
                              !is.na(tree_byPatient$AgeGroup) &
                              tree_byPatient$SampleType2 %in% c("Healthy", "RSV-I", "RSV-U")),],
       aes(x = SampleType2, y = meanDist_from_germline, colour = AgeGroup)) +
  geom_boxplot() + geom_point(position = position_jitterdodge()) + cowplot::theme_cowplot() +
  facet_grid(~ CSR) + xlab("") + ylab("mean distance from germline") + ylim(0, 0.28) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)),
  ncol = 1
)

```


## Statistical test: CSR event count

only shown those with p < 0.05:

```{r}
nEvent_test <- apply(combn(c("M", "G3", "G1", "A1", "G2", "G4", "E", "A2"), 2), 
                     MARGIN = 2, function(x){
  o <- lapply(1, function(y){#c("Young", "Middle", "Old"), function(y){
    lm_tb <- tree_byPatient[which(tree_byPatient$startIsotype == x[1] &
                                    tree_byPatient$endIsotype == x[2]), ]
#                                    tree_byPatient$AgeGroup == y), ]
    colnames(lm_tb)[which(colnames(lm_tb) == "SampleType2")] <- "xxx"
    
    # One-way ANOVA
    if(nrow(lm_tb) > 0 & length(unique(lm_tb$xxx)) > 1 &
       sum(grepl("^Healthy", lm_tb$xxx)) > 0){
      onewayanova <- try( suppressWarnings(
        aov(
          as.formula( paste("nClone_event", "xxx", sep = " ~ ")), data = lm_tb
        )), silent = TRUE)
      if( "aov" %in% class(onewayanova) ){
        # Do Post-hoc test (Dunnett)
        p <- summary(onewayanova)[[1]][1,5]
        if( is.null(p) ) posthoc <- data.frame() else {
          posthoc <- glht(onewayanova, linfct = mcp(xxx = "Dunnett"))
          posthoc <- summary(posthoc)
          signif_comparison <- 1:length(posthoc$test$pvalues)#which(posthoc$test$pvalues < 0.05)
          if( length(signif_comparison) > 0){
            posthoc <- lapply(signif_comparison, function(y){
              data.frame(startIsotype = x[1], endIsotype = x[2],
                         SampleType2 = unlist(strsplit(names(posthoc$test$coefficients)[y], split = " - "))[1],
                         pval = round(posthoc$test$pvalues[y], 4))
            })
            posthoc <- do.call("rbind", posthoc)
          } else posthoc <- data.frame()
        }
      } else return(data.frame())
    } else return(data.frame())
  })
  do.call("rbind", o)
})
nEvent_test <- do.call("rbind", nEvent_test)
rownames(nEvent_test) <- NULL
nEvent_test <- nEvent_test[nEvent_test$pval < 0.05, ]
nEvent_test
```

```{r, fig.width=7, fig.height=11}
# plot
tree_byPatient$CSR <- apply(tree_byPatient[, c("startIsotype", "endIsotype")],
                            MARGIN = 1, function(x) paste(x[1], x[2], sep = " to "))
nEvent_test$CSR <- apply(nEvent_test[, c("startIsotype", "endIsotype")],
                         MARGIN = 1, function(x) paste(x[1], x[2], sep = " to "))
tree_byPatient <- merge(tree_byPatient, 
                        nEvent_test[, c("SampleType2", "CSR", "pval")],
                        by = c("SampleType2", "CSR"), all.x = TRUE,
                        sort = FALSE)
colnames(tree_byPatient)[ncol(tree_byPatient)] <- "p_nEvent"
tree_byPatient$signif_nEvent <- (tree_byPatient$p_nEvent < 0.05)
tree_byPatient[is.na(tree_byPatient$signif_nEvent), "signif_nEvent"] <- FALSE

# plot those with significant differences in at least 1 sample types
ggplot(tree_byPatient[which(tree_byPatient$CSR %in% nEvent_test$CSR),],
       aes(x = SampleType2, y = nClone_event, colour = signif_nEvent)) +
  geom_boxplot() + geom_point() + cowplot::theme_cowplot() +
  facet_wrap( ~ CSR, ncol = 4, scales = "free_x") + xlab("") + #ylab("% clones with CSR events") +
  scale_y_continuous(labels = scales::percent, name="% clones with CSR events") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
  scale_colour_manual(values = c("TRUE" = "red", "FALSE" = "grey"), name = "p < 0.05")
```

Split by age group. Model $Y = \text{agegroup}\ *\ \text{SampleType} $ and Tukey's post-hoc test.

```{r}
# evaluate statistical significance
nEvent_test <- apply(combn(c("M", "G3", "G1", "A1", "G2", "G4", "E", "A2"), 2), 
                     MARGIN = 2, function(x){
  o <- lapply(1, function(y){#c("Young", "Middle", "Old"), function(y){
    lm_tb <- tree_byPatient[which(tree_byPatient$startIsotype == x[1] &
                                    tree_byPatient$endIsotype == x[2]), ]
#                                    tree_byPatient$AgeGroup == y), ]
    # colnames(lm_tb)[which(colnames(lm_tb) == "SampleType2")] <- "xxx"
    lm_tb$AgeSample <- interaction(lm_tb$AgeGroup, lm_tb$SampleType2)
    # Two-way ANOVA
    if(nrow(lm_tb) > 0 & length(unique(lm_tb$SampleType2)) > 1 &
       sum(grepl("^Healthy", lm_tb$SampleType2)) > 0){
      twowayanova <- try( suppressWarnings(
        aov(
          as.formula( paste("nClone_event ~ AgeSample - 1")), data = lm_tb
        )), silent = TRUE)
      if( "aov" %in% class(twowayanova) ){
        # Do Post-hoc test (Dunnett)
        p <- summary(twowayanova)[[1]][1,5]
        if( is.null(p) ) posthoc <- data.frame() else {
          Tukey <- table(lm_tb$AgeSample)
          Tukey <- contrMat(Tukey[Tukey > 0], "Tukey")
          # comparison of either sample age group & any sample type vs Healthy, or
          # same sample type & different age groups
          Tukey <- list(
            Tukey[which(grepl("Healthy", rownames(Tukey)) & stringr::str_count(rownames(Tukey), "<=50") != 1), ],
            Tukey[which(sapply(stringr::str_extract_all(rownames(Tukey), "\\.([A-Za-z\\-]*)"), 
                               function(z) z[1] ==z[2])), ]
          )
          Tukey <- do.call("rbind", Tukey)
          posthoc <- glht(twowayanova, linfct = Tukey)
          posthoc <- summary(posthoc)
          signif_comparison <- 1:length(posthoc$test$pvalues)#which(posthoc$test$pvalues < 0.05)
          if( length(signif_comparison) > 0){
            posthoc <- lapply(signif_comparison, function(y){
              data.frame(startIsotype = x[1], endIsotype = x[2],
                         group1 = unlist(strsplit(names(posthoc$test$coefficients)[y], split = " - "))[1],
                         group2 = unlist(strsplit(names(posthoc$test$coefficients)[y], split = " - "))[2],
                         pval = round(posthoc$test$pvalues[y], 4))
            })
            posthoc <- do.call("rbind", posthoc)
          } else posthoc <- data.frame()
        }
      } else return(data.frame())
    } else return(data.frame())
  })
  do.call("rbind", o)
})
nEvent_test <- do.call("rbind", nEvent_test)
rownames(nEvent_test) <- NULL
#nEvent_test <- nEvent_test
nEvent_test[nEvent_test$pval < 0.05, ]
```

```{r, fig.width=11, fig.height=8}
nEvent_test$CSR <- apply(nEvent_test[, c("startIsotype", "endIsotype")],
                         MARGIN = 1, function(x) paste(x[1], x[2], sep = " to "))
# plot those with significant differences in at least 1 sample type
g1 <- ggplot(tree_byPatient[which(tree_byPatient$CSR %in% nEvent_test[nEvent_test$pval < 0.05, "CSR"] & 
                              # grepl("^M", tree_byPatient$CSR) & 
                              !is.na(tree_byPatient$AgeGroup) &
                              tree_byPatient$SampleType2 %in% c("Healthy", "CV19", "EBOV (White)")),],
       aes(x = SampleType2, y = nClone_event, colour = AgeGroup)) +
  geom_boxplot() + geom_point(position = position_jitterdodge()) + cowplot::theme_cowplot() +
  facet_grid(~ CSR) + xlab("") + ylab("% clones with CSR events") + ylim(0, 0.1) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10), legend.position = "bottom")

g2 <- ggplot(tree_byPatient[which(tree_byPatient$CSR %in% nEvent_test[nEvent_test$pval < 0.05, "CSR"] & 
                              # grepl("^M", tree_byPatient$CSR) & 
                              !is.na(tree_byPatient$AgeGroup) &
                              tree_byPatient$SampleType2 %in% c("Healthy", "RSV-I", "RSV-U")),],
       aes(x = SampleType2, y = nClone_event, colour = AgeGroup)) +
  geom_boxplot() + geom_point(position = position_jitterdodge()) + cowplot::theme_cowplot() +
  facet_grid(~ CSR) + xlab("") + ylab("% clones with CSR events") + ylim(0, 0.1) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10), legend.position = "bottom")

cowplot::plot_grid(g1 + ggtitle("CV / EBOV"), g2 + ggtitle("RSV"), ncol = 1)
```

Differences driven by Age $\leq50$ samples? But probably simply because more data points?