---
title: "Mutational levels"
author: "Joseph Ng"
date: "6 February 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = "MutationalLevels_plots/",
                      dev = "svg")
library(plyr)
library(ggplot2)
library(multcomp)
#library(shazam)

```

Analysis of mutational levels of BCR clones by using tree-based distances, in repertoires collected from individuals with different immunological states.

```{r loaddata, include=FALSE}
test_data <- read.table('/media/josefng//My_Passport/CovidRepertoire/AllCV+AllHC+AllRSV+AllEB_wo10XYF.txt', header = TRUE, sep = "\t",
                      stringsAsFactors = FALSE)
test_data <- test_data[ which(grepl("^IGH", test_data$Vfamily)), ]

# set up age group
test_data$AgeGroup <- test_data$Age
test_data$AgeGroup[ which(as.numeric(test_data$Age) <= 50 )] <-  "<=50"
test_data$AgeGroup[ which(as.numeric(test_data$Age) >= 60 )] <-  ">=60"
test_data$AgeGroup[ which(as.numeric(test_data$Age) > 50 & as.numeric(test_data$Age < 60))] <- NA
#test_data$AgeGroup[ which(as.numeric(test_data$Age) > 65 )] <-  "Old"
test_data$AgeGroup[ which(test_data$Age == "nd")] <- NA
test_data$AgeGroup[ which(test_data$Age == "Young")] <- "<=50"
test_data$AgeGroup[ which(test_data$Age == "Old")] <- ">=60"
colnames(test_data)[ncol(test_data)] <- "AgeGroup"

# group RSV Day 0 as Healthy in a new SampleType definition
test_data$SampleType2 <- test_data$SampleType
test_data$SampleType2[which(test_data$SampleType %in% c("Infected", "Uninfected", "YF") & test_data$TimePoint == "0")] <- "Healthy"
test_data$SampleType2[which(test_data$SampleType %in% c("COVID") & test_data$TimePoint == "FU")] <- "COVID-19Recovered"
test_data$SampleType2[which(test_data$SampleType %in% c("COVID") & test_data$TimePoint != "FU")] <- "COVID-19"
test_data$SampleType2[which(test_data$SampleType %in% c("Infected") & test_data$TimePoint == "10")] <- "RSV-I"
test_data$SampleType2[which(test_data$SampleType %in% c("Uninfected") & test_data$TimePoint == "10")] <- "RSV-U"
test_data$SampleType2[which(test_data$SampleType %in% c("EB"))] <- "Ebola"
test_data$SampleType2[which(test_data$SampleType %in% c("YF") & test_data$TimePoint == "28")] <- "YFVD28"

# MutationalStatus
test_data$MutationalStatus <- sapply(test_data$V.REGION.identity.., function(x){
  if(as.numeric(x) < 99) return("Mutated") else if (as.numeric(x) >= 99) return("Unmutated")
  else return(NA)
})

# CellType
test_data$CellType <- test_data$Subclass
test_data$CellType[which(test_data$Subclass == "IgM" & test_data$MutationalStatus == "Unmutated" & test_data$NumInClone < 3)] <- "Naive"
test_data$CellType[which(test_data$Subclass == "IgM" & test_data$MutationalStatus %in% c("Unmutated", "Mutated") & test_data$NumInClone >= 3)] <- "IgM memory"

# CloneGroup
test_data$CloneGroup <- sapply(test_data$NumInClone, function(x){
  if(is.na(x)) return(NA)
  if(x == 1) return("Unique")
  else if(x == 2) return("2")
  else if(x == 3) return("3")
  else if(x %in% c(4, 5)) return("4&5")
  else if(x %in% 6:9) return("6 to 9")
  return(">10")
})
```

# Distance-from-germline distribution of trees

```{r eval=FALSE}
all_trees <- c(
  list.files(path = '/media/josefng//My_Passport/CovidRepertoire/CovidHB_alltrees/all_trees', full.names = TRUE),
  list.files(path = '/media/josefng//My_Passport/CovidRepertoire/otherReps_alltrees/newTrees', full.names = TRUE)
)
distFromGermline <- lapply(all_trees, function(ts){
  tree <- ape::read.tree(ts)
  if( length(tree$tip.label) < 7 ) return(NULL)
  root <- tree$tip.label[ which(grepl("^IG.V", tree$tip.label)) ]
  dists <- igraph::distances(alakazam::phyloToGraph(tree, germline = root),
                             v = root, 
                             to = tree$tip.label[which(tree$tip.label != root)])
  o <- data.frame(t(dists))
  colnames(o) <- "distFromGermline"
  o$SeqID <- rownames(o)
  rownames(o) <- NULL
  o[, c("SeqID", "distFromGermline")]
})
distFromGermline <- distFromGermline[ sapply( distFromGermline, 
                                              function(x) !is.null(x)) ]
distFromGermline <- do.call( "rbind", distFromGermline )
saveRDS(distFromGermline, 'allreps_distFromGermline.rds')
```

```{r fig.width=11, fig.height=10}
distFromGermline <- readRDS('allreps_distFromGermline.rds')
# match sample/patient IDs, UseAsRef, V gene Subclass back to here:
dist_metadata <- test_data[, c("Seq_ID", "SampleName", "PatientID", "Subclass",
                               "V.GENE.and.allele", "V.REGION.identity..",
                               "UseAsRef", "CloneID", "Class", "SampleType",
                               "AgeGroup", "SampleType2", "TimePoint")]
colnames(dist_metadata)[3] <- "PatientId"
dist_metadata$CloneID <- apply(dist_metadata[, c("PatientId", "CloneID")],
                               MARGIN = 1, function(x) paste(x[1], x[2], 
                                                             sep = "_"))
dist_metadata$CloneID <- gsub(" ", "", dist_metadata$CloneID)
dist_metadata$Vmut <- 100 - dist_metadata$V.REGION.identity..
distFromGermline <- merge(dist_metadata, distFromGermline, by.x = "Seq_ID",
                          by.y = "SeqID", sort = FALSE, all.y = TRUE)
distFromGermline <- distFromGermline[which(!is.na(distFromGermline$SampleName)), ]
colnames(distFromGermline)[ncol(distFromGermline)] <- "dist"
distFromGermline$Vmut_bin <- cut(distFromGermline$Vmut,
                                 include.lowest = TRUE, breaks = unique(quantile(distFromGermline$Vmut, probs = seq(0, 1, length.out = 21))))
distFromGermline$dist_bin <- ggplot2::cut_number(distFromGermline$dist, n = 20)
saveRDS(distFromGermline, 'allreps_distFromGermline_annotated.rds')

# count number of seqs in each bin and order the clones by median dist
distFromGermline_count <- split(distFromGermline, f = distFromGermline$SampleType2)
distFromGermline_count <- lapply(names(distFromGermline_count), function(x){
  tb <- distFromGermline_count[[x]]
  tb <- ddply(tb, .variables = c("CloneID", "dist_bin"), nrow, .drop = FALSE)
  tb$SampleType2 <- x
  tb
})
distFromGermline_count <- do.call("rbind", distFromGermline_count)
colnames( distFromGermline_count )[ ncol(distFromGermline_count)-1 ] <- "count" 
distFromGermline_count <- merge(distFromGermline_count,
                                ddply(distFromGermline,
                                      .variables = c("PatientId", "CloneID",
                                                     "AgeGroup", "SampleType2"),
                                      nrow))
colnames( distFromGermline_count )[ ncol(distFromGermline_count) ] <- "total" 
distFromGermline_count$perc <- distFromGermline_count$count / distFromGermline_count$total

distFromGermline_median <- ddply(distFromGermline,
                                 .variables = c("SampleType2", "PatientId", "CloneID"),
                                 summarise,
                                 dist_median = median(dist))
distFromGermline_median <- distFromGermline_median[ order(distFromGermline_median$dist_median), ]
distFromGermline_count$CloneID <- factor(distFromGermline_count$CloneID,
                                         levels = unique(distFromGermline_median$CloneID))
distFromGermline_count$SampleType2 <- factor(distFromGermline_count$SampleType2,
                                             levels = c("Healthy", "COVID-19", "COVID-19Recovered",
                                                        "Ebola", "RSV-I", "RSV-U", "YFVD28"),
                                             labels = c("Healthy", "CV19", "CV19-Recovered",
                                                        "EBOV", "RSV-I", "RSV-U", "YFVD28"))

plotDistGermlineHeatmap <- function(tb, title)
{
  ggplot( tb, aes(x = dist_bin, y = CloneID, fill = perc)) + 
    xlab("distance from germline") + geom_raster() + 
    scale_fill_viridis_c(name = "% sequences", labels = scales::percent) +
    ylab("Clones") + cowplot::theme_cowplot() + 
    theme(axis.text = element_blank(), axis.ticks = element_blank()) +
    facet_wrap(SampleType2 + AgeGroup ~ PatientId, scales = "free_y") +
    ggtitle(title)
}
plotDistGermlineHeatmap(
  distFromGermline_count[which(distFromGermline_count$SampleType2 == "Healthy"), ],
  "all clones - Healthy"
)
```

\newpage

```{r fig.width=11, fig.height=10}
plotDistGermlineHeatmap(
  distFromGermline_count[which(distFromGermline_count$SampleType2 %in%
                                         c("CV19", "CV19-Recovered")), ],
  "all clones - CV19"
)
```

\newpage

```{r fig.width=11, fig.height=10}
plotDistGermlineHeatmap(
  distFromGermline_count[which(distFromGermline_count$SampleType2 %in%
                                         c("RSV-I", "RSV-U", "EBOV")), ],
  "all clones - RSV + EBOV"
)

```

\newpage

These heatmaps can be transformed into a curve by taking the median distance-from-germline for each clone, and calculating the area-under-curve (AUC) which gives a metric per patient/donor for comparison. This AUC metric quanties **"Germline likeness"** of a repertoire in general. This is akin to sequence similarity of germline, but our metric quantifiers a given *repertoire* rather than a specific sequence.

```{r }
getDistPercentile <- function(dists, percentiles)
{
  sapply(dists, function(d){
    min(which(d < percentiles))/100
  })
}

# for each clone median / lower quartile / upper quartile of the 
# dist-from-gerline of sequences in the clone
distFromGermline_median <- ddply(distFromGermline,
                                 .variables = c("PatientId", "CloneID",
                                                "SampleType2", "AgeGroup"),
                                 summarise,
                                 dist_median = median(dist),
                                 dist_lowerq = quantile(dist, probs = 0.25),
                                 dist_upperq = quantile(dist, probs = 0.75))
# convert into percentiles
distFromGermline_median$dist_median <- getDistPercentile(
  distFromGermline_median$dist_median,
  quantile(distFromGermline$dist, probs = seq(0.01, 1, by = 0.01))
)
distFromGermline_median$dist_lowerq <- getDistPercentile(
  distFromGermline_median$dist_lowerq,
  quantile(distFromGermline$dist, probs = seq(0.01, 1, by = 0.01))
)
distFromGermline_median$dist_upperq <- getDistPercentile(
  distFromGermline_median$dist_upperq,
  quantile(distFromGermline$dist, probs = seq(0.01, 1, by = 0.01))
)
distFromGermline_median <- split(distFromGermline_median, 
                                 f = list(distFromGermline_median$SampleType2,
                                          distFromGermline_median$PatientId), 
                                 drop = TRUE)
# order the clone by median(dist-from-germline) and calculate percentile
distFromGermline_median <- lapply(distFromGermline_median, function(tb){
  tb <- tb[order(tb$dist_median), ]
  tb$clone_order <- (1:nrow(tb)) / nrow(tb)
  tb
})
distFromGermline_median <- do.call("rbind", distFromGermline_median)
distFromGermline_median$SampleType2 <- factor(distFromGermline_median$SampleType2,
                                          levels = c("Healthy", "COVID-19",
                                                     "COVID-19Recovered",
                                                     "Ebola", "RSV-I", "RSV-U",
                                                     "YFVD28"),
                                          labels = c("Healthy", "CV19", "CV19-Recovered",
                                                     "EBOV", "RSV-I", "RSV-U", "YFVD28"))

# clone percentile vs dist-from-germline percentile curves
g1 <- ggplot(distFromGermline_median[which(!is.na(distFromGermline_median$SampleType2)), ], 
       aes(x = dist_median, y = clone_order, 
           group = interaction(SampleType2, PatientId), 
           colour = SampleType2)) + geom_line() + 
  scale_colour_brewer(type = "qual", drop = FALSE) +
  xlab("relative dist-from-germline") + ylab("clone percentile") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  cowplot::theme_cowplot() + facet_wrap(~ SampleType2) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 90, hjust = 1))

# calculate AUC
getAUC <- function(tb)
{
  dDist <- c(diff(tb$dist_median), 0)
  dClone <- c(diff(tb$clone_order), 0)
  sum(tb$clone_order * dDist) + sum(dDist * dClone)/2
}
distFromGermline_AUC <- unique(distFromGermline_count[, c("PatientId", "SampleType2", "AgeGroup")])
distFromGermline_median <- split(distFromGermline_median, 
                                 f = list(distFromGermline_median$SampleType2,
                                          distFromGermline_median$PatientId), 
                                 drop = TRUE)
distFromGermline_AUC$AUC <- apply(
  distFromGermline_AUC, MARGIN = 1, 
  function(x) getAUC(distFromGermline_median[[paste0(x[2], ".", x[1])]])
)
distFromGermline_AUC$SampleType2 <- factor(distFromGermline_AUC$SampleType2,
                                          levels = c("Healthy", "CV19",
                                                     "CV19-Recovered",
                                                     "EBOV", "RSV-I", "RSV-U",
                                                     "YFVD28"))
distFromGermline_AUC <- distFromGermline_AUC[which(!is.na(distFromGermline_AUC$SampleType2)), ]

# statistical comparisons
distFromGermline_AUC <- merge(
  distFromGermline_AUC,
  ggpubr::compare_means(AUC ~ SampleType2, data = distFromGermline_AUC, 
                        hide.ns = TRUE, ref.group = "Healthy", 
                        label = "p.signif")[, c("group2", "p.adj")],
  by.x = "SampleType2", by.y = "group2", 
  all.x = TRUE, all.y = FALSE, sort = FALSE
)
distFromGermline_AUC$sig <- ifelse(is.na(distFromGermline_AUC$p.adj), FALSE,
                                   (distFromGermline_AUC$p.adj < 0.05))
g2 <- ggplot(distFromGermline_AUC, aes(x = SampleType2, y = AUC, colour = sig)) + 
  geom_hline(yintercept = median(distFromGermline_AUC[ which(distFromGermline_AUC$SampleType == "Healthy"), "AUC"],
                                 na.rm = TRUE), linetype = "dashed") +
  geom_boxplot(outlier.shape = NA) + geom_point() +
  scale_colour_manual(values = c("TRUE" = "red", "FALSE" = "grey"), 
                      name = "FDR < 0.05") +
  cowplot::theme_cowplot() + xlab("") + ylab("Germline likeness") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        plot.margin = margin(10, 2, 2, 2), legend.position = "bottom")

cowplot::plot_grid(g1, g2,
                   nrow = 1, rel_widths = c(2,1))
saveRDS(distFromGermline_AUC, "allreps_distFromGermline_AUC.rds")
```

Separate by age. Two-way ANOVA (Age & SampleType) and Tukey post-hoc.

```{r, fig.width=5, fig.height=4}
# evaluate statistical significance
distFromGermline_AUC$AgeSample <- interaction(distFromGermline_AUC$AgeGroup,
                                              distFromGermline_AUC$SampleType2)

# Tukey's test contrast matrix
Tukey <- table(distFromGermline_AUC$AgeSample)
Tukey <- contrMat(Tukey[Tukey > 0], "Tukey")
# comparison of either sample age group & any sample type vs Healthy, or
# same sample type & different age groups
Tukey <- list(
  Tukey[which(grepl("Healthy", rownames(Tukey)) & 
                stringr::str_count(rownames(Tukey), "<=50") != 1), ],
  Tukey[which(sapply(stringr::str_extract_all(rownames(Tukey), "\\.([A-Za-z\\-]*)"), 
                     function(z) z[1] ==z[2])), ]
)
Tukey <- do.call("rbind", Tukey)
summary(
  glht(
    aov(as.formula( paste("AUC ~ AgeSample - 1")), 
        data = distFromGermline_AUC), linfct = Tukey
  )
)

ggplot(distFromGermline_AUC,#[!is.na(distFromGermline_AUC$AgeGroup), ], 
       aes(x = SampleType2, y = AUC, colour = AgeGroup)) + 
    geom_boxplot(outlier.shape = NA) + geom_point(position = position_dodge(width = 0.7)) +
  scale_colour_manual(values = c("<=50" = "grey30", ">=60" = "grey70"), na.translate = TRUE, na.value = "lightgreen", name ="Age") +
    cowplot::theme_cowplot() + xlab("") + ylab("Germline likeness") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), 
          plot.margin = margin(10, 2, 2, 2), legend.position = "bottom")
```

\newpage

## Split by time-points

Is a concave curve (ie low AUC) a signature of convalescence and a convex curve indicative of B-cell response? 

$\rightarrow$ If we split by time points (& specify follow-up if applicable) do we see different patterns across the timecourse?

```{r}
distFromGermline_median <- ddply(distFromGermline,
                                 .variables = c("PatientId", "CloneID",
                                                "SampleType", "AgeGroup",
                                                "SampleName", "TimePoint"),
                                 summarise,
                                 dist_median = median(dist),
                                 dist_lowerq = quantile(dist, probs = 0.25),
                                 dist_upperq = quantile(dist, probs = 0.75))
distFromGermline_median$dist_median <- getDistPercentile(
  distFromGermline_median$dist_median,
  quantile(distFromGermline$dist, probs = seq(0.01, 1, by = 0.01))
)
distFromGermline_median$dist_lowerq <- getDistPercentile(
  distFromGermline_median$dist_lowerq,
  quantile(distFromGermline$dist, probs = seq(0.01, 1, by = 0.01))
)
distFromGermline_median$dist_upperq <- getDistPercentile(
  distFromGermline_median$dist_upperq,
  quantile(distFromGermline$dist, probs = seq(0.01, 1, by = 0.01))
)
distFromGermline_median <- split(distFromGermline_median, 
                                 f = distFromGermline_median$SampleName)
distFromGermline_median <- lapply(distFromGermline_median, function(tb){
  tb <- tb[order(tb$dist_median), ]
  tb$clone_order <- (1:nrow(tb)) / nrow(tb)
  tb
})

# calculate AUC
distFromGermline_AUC_all <- distFromGermline_AUC
distFromGermline_AUC <- unique(do.call("rbind", distFromGermline_median)[, c("PatientId", "SampleType", "AgeGroup", "SampleName", "TimePoint")])
distFromGermline_AUC$AUC <- sapply(
  distFromGermline_AUC$SampleName, 
  function(x) getAUC(distFromGermline_median[[x]])
)
distFromGermline_AUC$SampleType <- factor(distFromGermline_AUC$SampleType,
                                          levels = c("Healthy", "COVID", "EB",
                                                     "Infected",
                                                     "Uninfected", "YF"),
                                          labels = c("Healthy", "CV19", "EBOV",
                                                     "RSV-I", "RSV-U", "YFVD28"))
distFromGermline_AUC$TimePoint <- factor(distFromGermline_AUC$TimePoint,
                                         levels = c("0", "nd", "2", "3", "5",
                                                    "6", "7", "8", "9", "10",
                                                    "14", "22", "28", "A",
                                                    "B", "FU"), ordered = TRUE)
# remove the timepoints for the 1 RSV-I sample with multiple timepoints
distFromGermline_AUC <- distFromGermline_AUC[(1:nrow(distFromGermline_AUC))[-which(distFromGermline_AUC$PatientId == "YR001" & ! distFromGermline_AUC$TimePoint %in% c("0", "10"))], ]

ggplot(distFromGermline_AUC, aes(x = TimePoint, y = AUC, group = PatientId)) +
  geom_hline(yintercept = median(distFromGermline_AUC[ which(distFromGermline_AUC$SampleType == "Healthy"), "AUC"],
                                 na.rm = TRUE), linetype = "dashed") +
  geom_point(position = position_dodge()) + geom_line() + 
  cowplot::theme_cowplot() + xlab("Timepoint") + ylab("Germline likeness") +
  facet_wrap(~ SampleType, scales = "free_x")
saveRDS(distFromGermline_AUC, "allreps_distFromGermline_splitTimepoint_AUC.rds")
```

Showing only COVID-19 with side-by-side Healthy and Ebola boxplots:

```{r fig.width=3.5, fig.height=3}

cowplot::plot_grid(
  ggplot(distFromGermline_AUC_all[which(distFromGermline_AUC_all$SampleType2 %in% c("Healthy", "EBOV")),], aes(x = SampleType2, y = AUC)) + 
  geom_hline(yintercept = median(distFromGermline_AUC[ which(distFromGermline_AUC$SampleType == "Healthy"), "AUC"],
                                 na.rm = TRUE), linetype = "dashed") +
  geom_boxplot(outlier.shape = NA) + geom_point() + ylim(0, 1) +
  cowplot::theme_cowplot() + xlab("") + ylab("Germline likeness") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        plot.margin = margin(10, 2, 2, 2), legend.position = "bottom"),
  ggplot(distFromGermline_AUC[which(distFromGermline_AUC$SampleType == "CV19"), ], 
         aes(x = TimePoint, y = AUC, group = PatientId)) +
    geom_hline(yintercept = median(distFromGermline_AUC[ which(distFromGermline_AUC$SampleType == "Healthy"), "AUC"],
                                 na.rm = TRUE), linetype = "dashed") + ylim(0, 1) +
    geom_point(position = position_dodge()) + geom_line() + 
    cowplot::theme_cowplot() + xlab("Timepoint") + ylab("") +
    facet_wrap(~ SampleType, scales = "free_x") +
    theme(axis.line.y = element_blank(), axis.text.y = element_blank(), 
          axis.ticks.y = element_blank()),
  nrow = 1, align = "h", axis = "tb", rel_widths = c(2, 3)
)

# statistical test (Jonckheere-Terpstra test for trend)
print(DescTools::JonckheereTerpstraTest(AUC ~ TimePoint, data = distFromGermline_AUC[which(distFromGermline_AUC$SampleType == "CV19"), ]))
```

\newpage

## Split by isotypes

Differences between isotypes?

```{r fig.width=5.5, fig.height=6.5}
distFromGermline_median <- ddply(distFromGermline,
                                 .variables = c("PatientId", "CloneID",
                                                "SampleType2", "AgeGroup", "Class"),
                                 summarise,
                                 dist_median = median(dist),
                                 dist_lowerq = quantile(dist, probs = 0.25),
                                 dist_upperq = quantile(dist, probs = 0.75))
distFromGermline_median$dist_median <- getDistPercentile(
  distFromGermline_median$dist_median,
  quantile(distFromGermline$dist, probs = seq(0.01, 1, by = 0.01))
)
distFromGermline_median$dist_lowerq <- getDistPercentile(
  distFromGermline_median$dist_lowerq,
  quantile(distFromGermline$dist, probs = seq(0.01, 1, by = 0.01))
)
distFromGermline_median$dist_upperq <- getDistPercentile(
  distFromGermline_median$dist_upperq,
  quantile(distFromGermline$dist, probs = seq(0.01, 1, by = 0.01))
)
distFromGermline_median$PatientClass <- apply(distFromGermline_median[, c("SampleType2", "PatientId", "Class")], MARGIN = 1, function(x) paste(x[1], x[2], x[3], sep = "_"))
distFromGermline_median <- distFromGermline_median[which(!grepl("_NA$", distFromGermline_median$PatientClass)), ]
distFromGermline_median <- split(distFromGermline_median, 
                                 f = distFromGermline_median$PatientClass)
distFromGermline_median <- lapply(distFromGermline_median, function(tb){
  tb <- tb[order(tb$dist_median), ]
  tb$clone_order <- (1:nrow(tb)) / nrow(tb)
  tb
})

# calculate AUC
distFromGermline_AUC <- unique(do.call("rbind", distFromGermline_median)[, c("PatientId", "SampleType2", "AgeGroup", "Class", "PatientClass")])
distFromGermline_AUC$AUC <- sapply(
  distFromGermline_AUC$PatientClass, 
  function(x) getAUC(distFromGermline_median[[x]])
)
distFromGermline_AUC$SampleType2 <- factor(distFromGermline_AUC$SampleType2,
                                          levels = c("Healthy", "COVID-19",
                                                     "COVID-19Recovered",
                                                     "Ebola", "RSV-I", "RSV-U",
                                                     "YFVD28"),
                                          labels = c("Healthy", "CV19", "CV19-Recovered",
                                                     "EBOV", "RSV-I", "RSV-U", "YFVD28"))
distFromGermline_AUC <- distFromGermline_AUC[which(!is.na(distFromGermline_AUC$SampleType2)), ]
distFromGermline_AUC$Class <- factor(distFromGermline_AUC$Class,
                                     levels = c("M", "G", "A"),
                                     labels = c("IgM", "IgG", "IgA"))
distFromGermline_AUC <- distFromGermline_AUC[which(!is.na(distFromGermline_AUC$Class)), ]
distFromGermline_AUC <- merge(
  distFromGermline_AUC,
  ggpubr::compare_means(AUC ~ Class, data = distFromGermline_AUC, 
                        group.by = "SampleType2", label = "p.signif", hide.ns = TRUE,
                        ref.group = "IgM", label.y = 0.85)[, c("group2", "SampleType2", "p.adj")],
  by.x = c("SampleType2", "Class"), by.y = c("SampleType2", "group2"),
  all.x = TRUE, all.y = TRUE, sort = FALSE
)
distFromGermline_AUC$sig <- ifelse(is.na(distFromGermline_AUC$p.adj), FALSE,
                                   (distFromGermline_AUC$p.adj < 0.05))
ggplot(distFromGermline_AUC, aes(x = Class, y = AUC, colour = sig)) +
  geom_line(aes(group = PatientId), colour = "grey") + 
  geom_hline(yintercept = median(distFromGermline_AUC[ which(distFromGermline_AUC$SampleType == "Healthy" &
                                                               distFromGermline_AUC$Class == "IgM"), "AUC"],
                                 na.rm = TRUE), linetype = "dashed") + geom_boxplot() +
  geom_point(position = position_dodge()) + 
  scale_colour_manual(values = c("TRUE" = "red", "FALSE" = "grey"), 
                      name = "FDR < 0.05") + ylab("Germline likeness") +
  cowplot::theme_cowplot() + xlab("") + ylim(0, 1) +
  facet_wrap(~ SampleType2, scales = "free_x")
saveRDS(distFromGermline_AUC, "allreps_distFromGermline_splitIsotype_AUC.rds")

```

\newpage

## Split by subclasses

Differences between subclasses:

```{r fig.width=7, fig.height=8}
distFromGermline_median <- ddply(distFromGermline,
                                 .variables = c("PatientId", "CloneID",
                                                "SampleType2", "AgeGroup", "Subclass"),
                                 summarise,
                                 dist_median = median(dist),
                                 dist_lowerq = quantile(dist, probs = 0.25),
                                 dist_upperq = quantile(dist, probs = 0.75))
distFromGermline_median$dist_median <- getDistPercentile(
  distFromGermline_median$dist_median,
  quantile(distFromGermline$dist, probs = seq(0.01, 1, by = 0.01))
)
distFromGermline_median$dist_lowerq <- getDistPercentile(
  distFromGermline_median$dist_lowerq,
  quantile(distFromGermline$dist, probs = seq(0.01, 1, by = 0.01))
)
distFromGermline_median$dist_upperq <- getDistPercentile(
  distFromGermline_median$dist_upperq,
  quantile(distFromGermline$dist, probs = seq(0.01, 1, by = 0.01))
)
distFromGermline_median$PatientClass <- apply(distFromGermline_median[, c("SampleType2", "PatientId", "Subclass")], MARGIN = 1, function(x) paste(x[1], x[2], x[3], sep = "_"))
distFromGermline_median <- distFromGermline_median[which(!grepl("_NA$|[AG]$", distFromGermline_median$PatientClass)), ]
distFromGermline_median <- split(distFromGermline_median, 
                                 f = distFromGermline_median$PatientClass)
distFromGermline_median <- lapply(distFromGermline_median, function(tb){
  tb <- tb[order(tb$dist_median), ]
  tb$clone_order <- (1:nrow(tb)) / nrow(tb)
  tb
})

# calculate AUC
distFromGermline_AUC <- unique(do.call("rbind", distFromGermline_median)[, c("PatientId", "SampleType2", "AgeGroup", "Subclass", "PatientClass")])
distFromGermline_AUC$AUC <- sapply(
  distFromGermline_AUC$PatientClass, 
  function(x) getAUC(distFromGermline_median[[x]])
)
distFromGermline_AUC$SampleType2 <- factor(distFromGermline_AUC$SampleType2,
                                          levels = c("Healthy", "COVID-19",
                                                     "COVID-19Recovered",
                                                     "Ebola", "RSV-I", "RSV-U",
                                                     "YFVD28"),
                                          labels = c("Healthy", "CV19", "CV19-Recovered",
                                                     "EBOV", "RSV-I", "RSV-U", "YFVD28"))
distFromGermline_AUC <- distFromGermline_AUC[which(!is.na(distFromGermline_AUC$SampleType2)), ]
distFromGermline_AUC$Subclass <- factor(distFromGermline_AUC$Subclass,
                                     levels = c("IgM", "IgG1", "IgG2", "IgG3", "IgG4",
                                                "IgA1", "IgA2"))
distFromGermline_AUC <- distFromGermline_AUC[which(!is.na(distFromGermline_AUC$Subclass)), ]
distFromGermline_AUC <- merge(
  distFromGermline_AUC,
  ggpubr::compare_means(AUC ~ Subclass, data = distFromGermline_AUC, 
                        group.by = "SampleType2", label = "p.signif", hide.ns = TRUE,
                        ref.group = "IgM", label.y = 0.85)[, c("group2", "SampleType2", "p.adj")],
  by.x = c("SampleType2", "Subclass"), by.y = c("SampleType2", "group2"),
  all.x = TRUE, all.y = TRUE, sort = FALSE
)
distFromGermline_AUC$sig <- ifelse(is.na(distFromGermline_AUC$p.adj), FALSE,
                                   (distFromGermline_AUC$p.adj < 0.05))
ggplot(distFromGermline_AUC, aes(x = Subclass, y = AUC, colour = sig)) +
  geom_hline(yintercept = median(distFromGermline_AUC[ which(distFromGermline_AUC$SampleType == "Healthy" &
                                                               distFromGermline_AUC$Subclass == "IgM"), "AUC"],
                                 na.rm = TRUE), linetype = "dashed") + geom_boxplot() +
  geom_point(position = position_dodge()) + #geom_line(aes(group = PatientId)) + 
  scale_colour_manual(values = c("TRUE" = "red", "FALSE" = "grey"), name = "FDR < 0.05") +
  cowplot::theme_cowplot() + xlab("") + ylim(0, 1) +
  facet_wrap(~ SampleType2) + ylab("Germline likeness") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
saveRDS(distFromGermline_AUC, "allreps_distFromGermline_splitSubclass_AUC_all.rds")
```


## Is it the same if we consider sequence identity to the germline?

Use the V-region identity column from VQuest:

```{r fig.width=11, fig.height=10}
# count number of seqs in each bin and order the clones by median dist
distFromGermline_count <- split(distFromGermline, f = distFromGermline$SampleType2)
distFromGermline_count <- lapply(names(distFromGermline_count), function(x){
  tb <- distFromGermline_count[[x]]
  tb <- ddply(tb, .variables = c("CloneID", "Vmut_bin"), nrow, .drop = FALSE)
  tb$SampleType2 <- x
  tb
})
distFromGermline_count <- do.call("rbind", distFromGermline_count)
colnames( distFromGermline_count )[ ncol(distFromGermline_count)-1 ] <- "count" 
distFromGermline_count <- merge(distFromGermline_count,
                                ddply(distFromGermline,
                                      .variables = c("PatientId", "CloneID",
                                                     "AgeGroup", "SampleType2"),
                                      nrow))
colnames( distFromGermline_count )[ ncol(distFromGermline_count) ] <- "total" 
distFromGermline_count$perc <- distFromGermline_count$count / distFromGermline_count$total

distFromGermline_median <- ddply(distFromGermline,
                                 .variables = c("SampleType2", "PatientId", "CloneID"),
                                 summarise,
                                 dist_median = median(Vmut))
distFromGermline_median <- distFromGermline_median[ order(distFromGermline_median$dist_median), ]
distFromGermline_count$CloneID <- factor(distFromGermline_count$CloneID,
                                         levels = unique(distFromGermline_median$CloneID))
colnames(distFromGermline_count)[3] <- "dist_bin"

```

```{r }
# for each clone the median / lower quartile / upper quartile of the 
# (1 - % identity to V germline) distribution of all sequences in the clone
distFromGermline_median <- ddply(distFromGermline,
                                 .variables = c("PatientId", "CloneID",
                                                "SampleType2", "AgeGroup"),
                                 summarise,
                                 dist_median = median(Vmut),
                                 dist_lowerq = quantile(Vmut, probs = 0.25),
                                 dist_upperq = quantile(Vmut, probs = 0.75))
# convert to percentile
distFromGermline_median$dist_median <- getDistPercentile(
  distFromGermline_median$dist_median,
  quantile(distFromGermline$Vmut, probs = seq(0.01, 1, by = 0.01))
)
distFromGermline_median$dist_lowerq <- getDistPercentile(
  distFromGermline_median$dist_lowerq,
  quantile(distFromGermline$Vmut, probs = seq(0.01, 1, by = 0.01))
)
distFromGermline_median$dist_upperq <- getDistPercentile(
  distFromGermline_median$dist_upperq,
  quantile(distFromGermline$Vmut, probs = seq(0.01, 1, by = 0.01))
)

# order the clones by median % identity to germline and convert into a percentil
distFromGermline_median <- split(distFromGermline_median, 
                                 f = list(distFromGermline_median$SampleType2,
                                          distFromGermline_median$PatientId),
                                 drop = TRUE)
distFromGermline_median <- lapply(distFromGermline_median, function(tb){
  tb <- tb[order(tb$dist_median), ]
  tb$clone_order <- (1:nrow(tb)) / nrow(tb)
  tb
})
distFromGermline_median <- do.call("rbind", distFromGermline_median)
distFromGermline_median$SampleType2 <- factor(distFromGermline_median$SampleType2,
                                          levels = c("Healthy", "COVID-19",
                                                     "COVID-19Recovered",
                                                     "Ebola", "RSV-I", "RSV-U",
                                                     "YFVD28"),
                                          labels = c("Healthy", "CV19", "CV19-Recovered",
                                                     "EBOV", "RSV-I", "RSV-U", "YFVD28"))

# clone percentile vs V-identity percentile
g1 <- ggplot(distFromGermline_median[which(!is.na(distFromGermline_median$SampleType2)), ], 
       aes(x = dist_median, y = clone_order, 
           group = interaction(SampleType2, PatientId), 
           colour = SampleType2)) + geom_line() + 
  scale_colour_brewer(type = "qual", drop = FALSE) +
  xlab("1 - V-gene identity to germline(%)") + ylab("clone percentile") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  cowplot::theme_cowplot() + facet_wrap(~ SampleType2) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90, hjust = 1))

# calculate AUC
distFromGermline_count$SampleType2 <- factor(distFromGermline_count$SampleType2,
                                          levels = c("Healthy", "COVID-19",
                                                     "COVID-19Recovered",
                                                     "Ebola", "RSV-I", "RSV-U",
                                                     "YFVD28"),
                                          labels = c("Healthy", "CV19",
                                                     "CV19-Recovered",
                                                     "EBOV", "RSV-I", "RSV-U",
                                                     "YFVD28"))
distFromGermline_AUC <- unique(distFromGermline_count[, c("PatientId", "SampleType2", "AgeGroup")])
distFromGermline_median <- split(distFromGermline_median, 
                                 f = list(distFromGermline_median$SampleType2,
                                          distFromGermline_median$PatientId), 
                                 drop = TRUE)
distFromGermline_AUC$AUC <- apply(
  distFromGermline_AUC, MARGIN = 1, 
  function(x) getAUC(distFromGermline_median[[paste0(x[2], ".", x[1])]])
)
distFromGermline_AUC <- distFromGermline_AUC[which(!is.na(distFromGermline_AUC$SampleType2)), ]
distFromGermline_AUC <- merge(
  distFromGermline_AUC,
  ggpubr::compare_means(AUC ~ SampleType2, data = distFromGermline_AUC, hide.ns = TRUE, ref.group = "Healthy", label = "p.signif")[, c("group2", "p.adj")],
  by.x = "SampleType2", by.y = "group2", all.x = TRUE, all.y = FALSE, sort = FALSE
)
distFromGermline_AUC$sig <- ifelse(is.na(distFromGermline_AUC$p.adj), FALSE,
                                   (distFromGermline_AUC$p.adj < 0.05))
g2 <- ggplot(distFromGermline_AUC, aes(x = SampleType2, y = AUC, colour = sig)) + 
  geom_hline(yintercept = median(distFromGermline_AUC[ which(distFromGermline_AUC$SampleType == "Healthy"), "AUC"],
                                 na.rm = TRUE), linetype = "dashed") +
  geom_boxplot(outlier.shape = NA) + geom_point() +
  scale_colour_manual(values = c("TRUE" = "red", "FALSE" = "grey"), 
                      name = "FDR < 0.05") +
  cowplot::theme_cowplot() + xlab("") + ylab("Germline likeness") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        plot.margin = margin(10, 2, 2, 2), legend.position = "bottom")

cowplot::plot_grid(g1, g2,
                   nrow = 1, rel_widths = c(2,1))
saveRDS(distFromGermline_AUC, "allreps_Videntity_AUC.rds")
```

Same picture as in the dist-from-germline derived from trees.
